---
title: 'DSA 8020 R Lab 4: Model Selection and Model Checking'
author: "your name here, (the names of any collaborators)"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Savings rates in 50 countries

The savings data frame has 50 rows (countries) and 5 columns (variables):

1. \texttt{sr}: savings rate - personal saving divided by disposable income
*This variable will be used as the response*
2. \texttt{pop15}: percent population under age of 15
3. \texttt{pop75}: percent population over age of 75
4. \texttt{dpi}: per-capita disposable income in dollars
5. \texttt{ddpi}: percent growth rate of dpi

The data is averaged over the period 1960-1970. 



*Data Source:* Belsley, D., Kuh. E. and Welsch, R. (1980) *Regression Diagnostics* Wiley.


Load the dataset

**Code:**

```{r}
data(savings, package = "faraway")
head(savings)
```

1. Perform the best subset selection and select the "best" model using $R^{2}_{adj}$

**Code:**

```{r}

```

**Answer:**


2. Perform a stepwise selection using *AIC*

**Code:**

```{r}

```

**Answer:**

3. Perform a general linear F-test (with $\alpha = 0.1$) to choose between the full model (i.e., using the all 4 predictors) and the reduce model that inlcude \texttt{pop15}, \texttt{pop75}, and \texttt{ddpi} as the predictors 

**Code:**

```{r}

```

**Answer:**

4. Make a residual plot of the model selected by *AIC* and comment the model assumptions

**Code:**

```{r}

```

**Answer:**

5. Use both histogram and qqplot to examine the normality assumption on error 

**Code:**

```{r}

```

**Answer:**

6. Calculate the leverage values to check if there is any high leverage points (i.e., $h>\frac{2p}{n}$)

**Code:**

```{r}

```

**Answer:**

7. Compute jackknife residuals to identify outlier(s)

**Code:**

```{r}

```

**Answer:**

8. Identifying influential observations by computing DFFITS

**Code:**

```{r}

```

**Answer:**



