---
title: 'DSA 8020 R Lab 3: Multiple Linear Regression II'
author: "your name here, (the names of any collaborators)"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Percentage of Body Fat and Body Measurements

Age, weight, height, and 10 body circumference measurements are recorded for 252 men. Each man's percentage of body fat was accurately estimated by an underwater weighing technique.



*Data Source:* Johnson R. *Journal of Statistics Education* v.4, n.1 (1996)


## Load the dataset

**Code:**

```{r}
library(faraway)
data(fat)
head(fat)
```

For the purposes of this lab, we will use only the following variables for conducting data analysis:

1. $y$ \texttt{brozek}: Percent body fat using Brozek's equation
$$\frac{457}{\text{Density}} - 414.2$$

2. $x_{1}$ \texttt{age}: Age (yrs); 

3. $x_2$\texttt{weight}: Height (inches);

4. $x_3$ \texttt{height}: Height (inches);

5. $x_4$\texttt{chest}: Chest circumference (cm);

6. $x_5$ \texttt{abdom}: Abdomen circumference (cm) at the umbilicus and level with the iliac crest

**Code:**

You can use the code below to extract these variables

```{r}
vars <- c("brozek", "age", "weight", "height", "chest", "abdom")
data <- fat[, vars]
```

 
## Exploratory Data Analysis

### Numerical summary

1. Use \texttt{summary} commend to produce various numerical summmaries of each of the 6 variables under consideration  

**Code:**

```{r}

```

### Graphical summary

2. Make a boxplot for each variable 

**Code:**

```{r}

```

3. Briefly discuss the shape of the distribution of each variable

**Answer:**


4. Create a scatterplot matrix to explore the inter-dependence between these these variables 

**Code:**

```{r}

```

## General Linear F-Test

Suppose a researcher would like to compare between the "Full" model using all the 5 predictors and a "reduce" model where only $x_{2}$ (\texttt{age}) and $x_{5}$ (\texttt{abdom}) are used by performing a general linear F-test:

5. Write down the null and the alternative hypotheses

**Answer:**

6. Fit the full model and write down the fitted linear regression equation.

**Code:**

```{r}

```

**Answer:**

7. Fit the reduce model and write down the fitted linear regression equation.

**Code:**

```{r}

```

**Answer:**

8. Perform a general linear F-test and state the conclusion at $\alpha = 0.05$

**Code:**

```{r}

```

**Answer:**

## Prediction

9. Predict a future response for an individal with \texttt{age} $=54$, \texttt{weight} $=197$, \texttt{height} $=72.25$, \texttt{chest} $=105.375$, and \texttt{abdom} $= 99.325$. Construct a 95% prediction interval.

**Code:**

```{r}

```


**Answer:**

10. Construct a 95% confidence interval for the mean response of percent body fat
 with \texttt{age} $=54$, \texttt{weight} $=197$, \texttt{height} $=72.25$, \texttt{chest} $=105.375$, and \texttt{abdom} $= 99.325$.
 
## Multicollinearity


11. Make the scatterplot matrix and compute the correlation matrix for all 6 variables (including the response).

**Code:**

```{r}

```




12. Calculate VIF and briefly discuss your finding

**Code:**

```{r}

```

**Answer:**




