---
title: "DSA 8020 R Lab 5: Analysis of covariance and Non-linear Regression"
author: "your name here, (the names of any collaborators)"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Analysis of covariance: Salaries for Professors

The 2008-09 nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S. The data were collected as part of the on-going effort of the college's administration to monitor salary differences between male and female faculty members.

### Load the dataset

**Code:**

```{r}
library(carData)
data(Salaries)
head(Salaries)
```

**Description of the variables**

* \texttt{rank}: a factor with levels Assistant Professor ("AsstProf"); Associate Professor ("AssocProf"); Full Professor ("Prof")

* \texttt{discipline}: a factor with levels A ("theoretical" departments) or B ("applied" departments)

* \texttt{yrs.since.phd}: years since her/his PhD

* \texttt{sex}: a factor with levels "Female" and "Male"

* \texttt{salary}: nine-month salary, in dollars

### Exploratory Data Analysis

1. Identify the numerical variables and categorical variabes in this data set

**Answer:**

2. Summarize each varaible numerically and graphically, briefly describe your findings

**Code:**

```{r}

```

**Answer:**

3. Create a scatterplot matrix and briefly describe your findings

**Code:**

```{r}

```

**Answer:**

### Model Fitting

4. Fit a MLR with \texttt{yrs.since.phd}, \texttt{discipline}, \texttt{rank}, and \texttt{sex} as the predictors. Write down the fitted regression equations for each category  (e.g., Feamle, Assistant Professor, theoretical departments). There are 12 categories in total 

**Code:**

```{r}

```

**Answer:**

5. State the model assumptions in the previous regression model


**Answer:**

6. Now fit another MLR with \texttt{yrs.since.phd}, \texttt{discipline}, \texttt{sex} and their interactions. Write down the fitted regression equations for each category

**Code:**

```{r}

```

**Answer:**

## Non-linear Regression: An Simulated Example

Suppose the response $y$ depends on the predictor $t$ in the following form:
$$y = \alpha \exp(-\beta t) + \epsilon,$$
where $\epsilon \sim \mathrm{N}(0, \sigma^2)$, and the true $\alpha$, $\beta$, and $\sigma^2$
are 15, 2 and 0.16, respectively. First, let's simulate some data points from this nonlinear model:

**Code:**

```{r}
alpha = 15; beta = 2; sigma.sq = 0.09
n <- 50
t <- seq(0, 2, len = 50)
set.seed(123)
y <- alpha * exp(-beta * t) + rnorm(n, sd = sqrt(sigma.sq))
data <- data.frame(y = y, t = t)

plot(t, y, las = 1, pch = 16)
lines(t, alpha * exp(-beta * t), type = "l", col = "blue")
legend("topright", legend = c("Data", "True"), pch = c(16, NA), lty = c(NA, 1),
       col = c("black", "blue"), bty = "n")
```

7. Use \texttt{nls} function to obtain nonlinear least-squares estimates $\hat{\alpha}$, $\hat{\beta}$, and $\hat{\sigma}^2$. In order to use \texttt{nls} you would need to provide formula = y $\sim$ alpha * exp(-beta * t), start = list(alpha = alpha_0, beta = beta_0), where alpha_0 and  beta_0 are initial guesses of the parameters $\alpha$ and $\beta$


**Code:**

```{r}

```

**Answer:**

8. Write down the fitted equation and the $\hat{\sigma}^2$  

**Answer:**

9. Apply the natural log transformation to the simulated response and fit a simple linear regression. Back transform to get the fit on the original scale. 

**Code:**

```{r}

```

**Answer:**

10. Comaring the nonlinear regression method and the linear regression with log-transformed response, which method would you prefer in this example. Explain your answer

**Answer:**

