---
title: 'DSA 8020 R Lab 1: Simple Linear Regression'
author: "your name here, (the names of any collaborators)"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Leaning Tower of Pisa

The dataset \texttt{PisaTower.csv} provides the annual measurements of the lean (the difference
between where a point on the tower would be if the tower were straight and where it
actually is) from 1975 to 1987. We would like to characterize lean over time by fitting a simple linear regression.


## Load the dataset

**Code:**

```{r}

```


## Descriptive analysis

### Numerical summary

**Code:**

```{r}

```

### Graphical summary

**Code:**

```{r}

```


**Question:** Describe the direction, strength, and the form of the relationship.

**Answer:**

## Simple linear regression


1. Identify the response variable, the predictor variable, and the sample size.

**Answer:**

2. Fit a simple linear regression 

**Code:**

```{r}

```

3. Write down the fitted linear regression model 

**Answer:**

4. What is $\hat{\sigma}$, the estimate of $\sigma$

**Answer:**

5. Find a 95% confidence interval for $\beta_{1}$

**Code:**

```{r}

```

6. Test the following hypothesis: $H_{0}: \beta_{1} = 0$ vs. $H_{a} : \beta_{1} \neq 0$ with $\alpha = 0.05$

**Answer:**


7. Construct a 90% confidence interval for $\mathrm{E}[Lean]$ in year 1984

**Code:**

```{r}

```

8. Use residuals to check model assumptions 

**Code:**

```{r}

```

**Answer:**

9. Would it be a good idea to use the fitted linear regression equation to predict \texttt{lean} in year 2010? Explain your answer. 

**Answer:**
