---
title: 'DSA 8020 R Lab 2: Multiple Linear Regression I'
author: "your name here, (the names of any collaborators)"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Housing Values in Suburbs of Boston

The Boston housing data was collected in 1978, each of the 506 entries represent aggregated data about 14 features for homes from various suburbs in Boston, MA. 

*Data Source:* Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. **J. Environ. Economics and Management** 5, 81–102.


## Load the dataset

**Code:**

```{r}
library(MASS)
data(Boston)
head(Boston)
```

For the purposes of this lab, we will use only the following variables for conducting data analysis:

1. \texttt{medv}: median value of owner-occupied homes in $\$1000s$; 

2. \texttt{lstat}: lower status of the population (percent);

3. \texttt{rm}: average number of rooms per dwelling;

4. \texttt{crim}: per capita crime rate by town

**Code:**

You can use the code below to extract these variables

```{r}
vars <- c("medv", "lstat", "rm", "crim")
data <- Boston[, vars]
```

 
## Exploratory Data Analysis

### Numerical summary

1. Use \texttt{summary} commend to produce various numerical summmaries of each of the 4 variables under consideration  

**Code:**

```{r}

```

### Graphical summary

2. Make a boxplot for each variable 

**Code:**

```{r}

```

3. Briefly discuss the shape of the distribution of each variable

**Answer:**


4. Create a scatterplot matrix to explore the inter-dependence between these these variables 

**Code:**

```{r}

```

## Model Fitting 

Here we will use \texttt{medv} as the response and \texttt{lstat}, \texttt{rm}, \texttt{crim} as predictors.  

### Simple Linear Regression

5. Fit a simple linear regression

**Code:**

```{r}

```

6. Write down the fitted linear regression equation.

**Answer:**

### Multiple Linear Regression

7. Fit a multiple linear regression using all predictors

**Code:**

```{r}

```

8. Write down the fitted linear regression equation

**Answer:**

9. Perform an overall F-test, state the hypotheses, test statistic, p-value, decision, and
conclusion 

**Answer:**




