---
title: "DSA 8020 R Session 9: Randomized Complete Block Designs, Factorial Designs, and Split-Plot Designs"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 7
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## RCBD 

### Create the data set

```{r}
x <- c(52, 47, 44, 51, 42, 60, 55, 49, 52, 43, 56, 48, 45, 44, 38)
trt <- rep(c("A", "B", "C"), each = 5)
blk <- rep(1:5, 3)
dat <- data.frame(x = x, trt = trt, blk = as.factor(blk))
```

### Two-way ANOVA

```{r}
lm <- lm(x ~ trt + blk, data = dat)
anova(lm)
```

### One-way ANOVA

```{r}
lm2 <- lm(x ~ trt, data = dat)
anova(lm2)
```

### Interaction plot: assessing the additivity assumption

```{r}
interaction.plot(dat$trt, dat$blk, x, las = 1, col = 1:5)
```

## Factorial Design

### Create the data set

```{r}
y <- c(130, 74, 150, 159, 138, 168, 155, 180, 188, 126, 110, 160,
       34, 80, 136, 106, 174, 150, 40, 75, 122, 115, 120, 139,
       20, 82, 25, 58, 96, 82, 70, 58, 70, 45, 104, 60)
temp <- c(rep(15, 12), rep(70, 12), rep(125, 12))
material <- rep(c(1, 1, 2, 2, 3, 3), 6)
dat <- data.frame(cbind(y, temp, material))
dat$temp <- as.factor(dat$temp); dat$material <- as.factor(dat$material)

(meanA <- tapply(dat$y, dat$temp, mean))
(meanB <- tapply(dat$y, dat$material, mean))
(meanAB <- tapply(dat$y, list(dat$temp, dat$material), mean))
```

### Two-way ANOVA

```{r}
lm <- lm(y ~ temp * material, data = dat)
anova(lm)
```

### Interaction plot

```{r}
interaction.plot(dat$temp, dat$material, dat$y, las = 1)

par(mar = c(4.1, 4, 1, 0.8))
plot(rep(c(15, 70, 125), 3), c(meanAB), pch = 16, las = 1,
     xlab = "Temperature (degree F)", ylab = "Average life (hrs)")
lines(c(15, 70, 125), meanAB[, 1])
lines(c(15, 70, 125), meanAB[, 2], lty = 5)
lines(c(15, 70, 125), meanAB[, 3], lty = 3)
legend("topright", legend = paste("Material", 1:3), bty = "n",
lty = c(1, 5, 3)) 
```


## Split-plot design

This example is taken from Lukas Meier's ANOVA using R  [\color{blue}{[Link]}](https://stat.ethz.ch/~meier/teaching/anova/split-plot-designs.html)

Farmer John has eight plots of land. He randomly assign two fertilization "schemes" ("control" and "new") to the eight plots. In addition, each plot (the "whole-plot") is divided into four subplots ("split-plots"). In each subplot, four different strawberry varieties are randomized to the subplots. John is interested in the effect of fertilization scheme and strawberry variety on fruit mass.

### Read the data

```{r}
dat <- read.table("http://stat.ethz.ch/~meier/teaching/data/john.dat", header = TRUE)
dat[, "plot"] <- factor(dat[, "plot"])
str(dat)
```

### ANOVA

```{r,message=FALSE}
#install.packages("lmerTest")
library(lmerTest)
fit <- lmer(mass ~ fertilizer * variety + (1 | plot), data = dat)
anova(fit)
```

### Interaction plot

```{r}
with(dat, interaction.plot(x.factor = variety, trace.factor = fertilizer, response = mass, las = 1))
```


