---
title: 'DSA 8020 R Lab 7: Completely Randomized Designs'
author: "your name here, (the names of any collaborators)"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The dataset \texttt{PlantGrowth} contains results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions.

*Data Source:* Dobson, A. J. (1983) *An Introduction to Statistical Modelling*. London: Chapman and Hall.

Let's load the data first:

**Code:**

```{r}
data(PlantGrowth)
str(PlantGrowth)
```

1. Compute the treatment means and standard deviations 

**Code:**

```{r}

```

2. Make side-by-side boxplots by treatment 

**Code:**

```{r}

```

3. Write down the effets model, explain each term in the model (including the model assumptions regarding the random error)

**Answer:**

4. Perform an overall F-test using ANOVA, state the hypotheses, p-value, decision, and
conclusion 

**Code:**

```{r}

```

**Answer:**



5. Conduct pairwise comparisons using Tukey's HSD  procedure

**Code:**

```{r}

```

**Answer:**

6. Use qqplot to examine the normality assumption on error 

**Code:**

```{r}

```

**Answer:**

7. Make a residual plot to assess equal variance assmuption

**Code:**

```{r}

```

**Answer:**

8. Perform a Levene’s test for equal variance

**Code:**

```{r}

```

**Answer:**





