---
title: "DSA 8020 R Session 5: Multiple Linear Regression IV"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 4
    fig_width: 6
    fig_height: 5
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression with Both Quantitative and Qualitative Predictors 

### Salaries for Professors Data Set

The 2008-09 nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S. The data were collected as part of the on-going effort of the college's administration to monitor salary differences between male and female faculty members.

### Load the data

```{r}
library(carData)
data(Salaries)
head(Salaries)
```

### Summazrize the data

```{r}
summary(Salaries)
boxplot(salary ~ sex, data = Salaries, las = 1)
boxplot(salary ~ discipline + sex, data = Salaries, las = 1)
boxplot(salary ~ rank, data = Salaries, las = 1)
# Cross tabulation
xtabs(~ sex + rank + discipline, data = Salaries)
# Plot salary vs. yrs since Ph.D. by gender using ggplot
library(ggplot2)
(plot1 <- ggplot(aes(x = yrs.since.phd, y = salary), data = Salaries))
(plot2 <- plot1 + geom_point(size = 0.5, colour = "red"))
(plot3 <- plot2 + facet_grid(~ sex))
(plot4 <- plot3 + geom_smooth(method = "lm"))
```

### Model fitting 

#### Model 1: A MLR with \texttt{yrs.since.phd} (numerical predictor), discipline, rank, and sex (categorical predictors)

```{r}
m1 <- lm(salary ~ discipline + rank + sex + yrs.since.phd, data = Salaries)
X <- model.matrix(m1)
head(X)
summary(m1)
```

### Plot the results of model 1 fit

```{r, fig.align='left'}
attach(Salaries)
yr.range <- tapply(yrs.since.phd, list(discipline, sex, rank), range)
sex.col <- ifelse(sex == "Male", "blue", "red")
dis.col <- ifelse(discipline == "A", 16, 1)

beta0 <- m1$coefficients[1]
betaDisp <- m1$coefficients[2]
betaAssoc <- m1$coefficients[3]
betaProf <- m1$coefficients[4]
betaMale <- m1$coefficients[5]
beta1 <- m1$coefficients[6]

library(scales)
# Plot the model fits by rank
## Assist prof
assistant <- which(rank == "AsstProf")
plot(yrs.since.phd[assistant], salary[assistant], pch = dis.col[assistant], cex = 0.8,
     col = alpha(sex.col[assistant], 0.5), yaxt = "n", xlab = "Years since PhD",
     main = "9-month salary", ylab = "")
axis(2, at = seq(63000, 99000, len = 6), labels = paste(seq(63000, 99000, len = 6)/ 1000, "k"),
     las = 1)

segments(yr.range[[1]][1], beta0 + yr.range[[1]][1] * beta1,
         yr.range[[1]][2], beta0 + yr.range[[1]][2] * beta1, col = "red", lwd = 1.8)
segments(yr.range[[2]][1], beta0 + betaDisp + yr.range[[2]][1] * beta1,
         yr.range[[2]][2], beta0 + betaDisp + yr.range[[2]][2] * beta1,
         col = "red", lty = 2, lwd = 1.8)
segments(yr.range[[3]][1], beta0 + betaMale + yr.range[[3]][1] * beta1,
         yr.range[[3]][2], beta0 + betaMale + yr.range[[3]][2] * beta1,
         col = "blue", lwd = 1.8)
segments(yr.range[[4]][1], beta0 + betaDisp + betaMale + yr.range[[4]][1] * beta1,
         yr.range[[4]][2], beta0 + betaDisp + betaMale + yr.range[[4]][2] * beta1,
         col = "blue", lty = 2, lwd = 1.8)
```


```{r}
## Assoc prof
assoc <- which(rank == "AssocProf")
plot(yrs.since.phd[assoc], salary[assoc], pch = dis.col[assoc], cex = 0.8,
     col = alpha(sex.col[assoc], 0.5), yaxt = "n", xlab = "Years since PhD",
     main = "9-month salary", ylab = "")
axis(2, at = seq(62000, 127000, len = 6), labels = paste(seq(62000, 127000, len = 6)/ 1000, "k"),
     las = 1)

segments(yr.range[[5]][1], beta0 + betaAssoc + yr.range[[5]][1] * beta1,
         yr.range[[5]][2], beta0 + betaAssoc + yr.range[[5]][2] * beta1,
         col = "red", lwd = 1.8)
segments(yr.range[[6]][1], beta0 + betaDisp + betaAssoc + yr.range[[6]][1] * beta1,
         yr.range[[6]][2], beta0 + betaDisp + betaAssoc + yr.range[[6]][2] * beta1,
         col = "red", lty = 2, lwd = 1.8)
segments(yr.range[[7]][1], beta0 + betaAssoc + betaMale + yr.range[[7]][1] * beta1,
         yr.range[[7]][2], beta0 + betaAssoc + betaMale + yr.range[[7]][2] * beta1,
         col = "blue", lwd = 1.8)
segments(yr.range[[8]][1], beta0 + betaDisp + betaAssoc + betaMale + yr.range[[8]][1] * beta1,
         yr.range[[8]][2], beta0 + betaDisp + betaAssoc + betaMale + yr.range[[8]][2] * beta1,
         col = "blue", lty = 2, lwd = 1.8)
```

```{r}
## Full prof
prof <- which(rank == "Prof")
plot(yrs.since.phd[prof], salary[prof],
     pch = dis.col[prof], cex = 0.8,
     col = alpha(sex.col[prof], 0.5),
     yaxt = "n", xlab = "Years since PhD",
     main = "9-month salary", ylab = "")
axis(2, at = seq(57000, 232000, len = 6),
     labels = paste(seq(57000, 232000, len = 6)/ 1000, "k"),
     las = 1)
segments(yr.range[[9]][1], beta0 + betaProf + yr.range[[9]][1] * beta1,
         yr.range[[9]][2], beta0 + betaProf + yr.range[[9]][2] * beta1,
         col = "red", lwd = 1.8)
segments(yr.range[[10]][1], beta0 + betaDisp + betaProf + yr.range[[10]][1] * beta1,
         yr.range[[10]][2], beta0 + betaDisp + betaProf + yr.range[[10]][2] * beta1,
         col = "red", lty = 2, lwd = 1.8)
segments(yr.range[[11]][1], beta0 + betaProf + betaMale + yr.range[[11]][1] * beta1,
         yr.range[[11]][2], beta0 + betaProf + betaMale + yr.range[[11]][2] * beta1,
         col = "blue", lwd = 1.8)
segments(yr.range[[12]][1], beta0 + betaDisp + betaProf + betaMale + yr.range[[12]][1] * beta1,
         yr.range[[12]][2], beta0 + betaDisp + betaProf + betaMale + yr.range[[12]][2] * beta1,
         col = "blue", lty = 2, lwd = 1.8)
```

```{r}
## Using ggplot
plot <- ggplot(aes(x = yrs.since.phd, y = salary), data = Salaries)
plot <- plot + geom_point()
plot <- plot + facet_grid(~ sex + rank)
(plot <- plot + geom_smooth(method = "lm"))
```

#### Model 2: Another MLR where we include the *interaction* between \texttt{sex} and \texttt{yrs.since.phd}

```{r}
m2 <- lm(salary ~ sex * yrs.since.phd)
summary(m2)
coeff <- m2$coefficients
plot(yrs.since.phd, salary, las = 1, pch = 16, cex = 0.5, col = alpha(sex.col, 0.5),
     xlab = "Years since PhD", main = "9-month salary", ylab = "")
grid()
abline(coeff[1], coeff[3], col = "red")
abline(coeff[1] + coeff[2], coeff[3] + coeff[4], col = "blue")
legend("toplef", legend = c("Female", "Male"),
       pch = 16, col = c("red", "blue"), bty = "n")
```

#### Model 3: One more MLR where we include the *interaction* between \texttt{discipline} and \texttt{yrs.since.phd}

```{r}
m3 <- lm(salary ~ discipline * yrs.since.phd)
summary(m3)
coeff <- m3$coefficients
plot(yrs.since.phd, salary, las = 1, pch = dis.col, cex = 0.5, col = alpha(sex.col, 0.5),
     xlab = "Years since PhD", main = "9-month salary", ylab = "")
grid()
abline(coeff[1], coeff[3])
abline(coeff[1] + coeff[2], coeff[3] + coeff[4], lty = 2)
legend("toplef", legend = c("Female", "Male"),
       pch = 16, col = c("red", "blue"), bty = "n")
```

## Polynomial regression


### Housing values in suburbs of Boston

* Dependent variable: *medv*, the median value of owner-occupied homes (in thousands of dollars).

* Independent variable: *lstat* (percent of lower status of the population).


### Load and plot the data

```{r}
library(MASS)
data(Boston)
plot(Boston$lstat, Boston$medv, col = "gray", pch = 16,
     cex = 0.6, las = 1, xlab = "lower status of the population (percent)",
     ylab = "median value of owner-occupied homes")
## ggplot
plot <- ggplot(aes(x = lstat, y = medv), data = Boston)
(plot <- plot + geom_point(colour = "gray"))
```

### Plot the poylnomial regression fits

```{r}
plot(Boston$lstat, Boston$medv, col = "gray", pch = 16,
     cex = 0.6, las = 1, xlab = "lower status of the population (percent)", ylab = "median value of owner-occupied homes")
## SLR
m1 <- lm(medv ~ lstat, data = Boston)
abline(m1)
## 2nd order polynomial fit
m2 <- lm(medv ~ lstat + I(lstat^2), data = Boston)
lines(sort(Boston$lstat), m2$fitted.values[order(Boston$lstat)], col = "red")
## 3rd order polynomial fit
m3 <- lm(medv ~ lstat + I(lstat^2) + I(lstat^3), data = Boston)
lines(sort(Boston$lstat), m3$fitted.values[order(Boston$lstat)], col = "blue")

## Using ggplot
plot <- plot + geom_smooth(method = "lm", colour = "black", se = F)
plot <- plot + geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "red", se = F)
plot <- plot + geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3), colour = "blue", se = F)
plot
```

### Model selection

```{r}
anova(m2, m3)
## Use Orthogonal Polynomials
m2new <- lm(medv ~ poly(lstat, 2), data = Boston)
m3new <- lm(medv ~ poly(lstat, 3), data = Boston)
summary(m3new); summary(m3)
anova(m2new, m3new)
```


## Nonlinear Regression

### U.S. Population Example

```{r}
library(car)
plot(population ~ year, data = USPop, main = "U.S. population",
     ylim = c(0, 300),pch = "*", xlab = "Census year",
     ylab = "Population in millions", cex = 1.25, las = 1, col = "blue")
grid()
```

### Logistic growth curve

A logistic function is a symmetric S shape curve with equation:
$$f(x) = \frac{\phi_{1}}{1+\exp(-(x - \phi_{2})/\phi_{3})}$$
where
$\phi_{1}$ is the curve's maximum value;
$\phi_{2}$ is the curve's midpoint in $x$;
and $\phi_{3}$ is the "range" (or the inverse growth rate) of the curve.

One typical application of the logistic equation is to model population growth. 

```{r}
# phi_1 = 10; phi_2 = 4/3, phi_3 = 1
curve(10 / (1 + exp(-(x - 4/3))), from = -8, to = 10, main = "Logistic growth curve",
      las = 1, xlab = "", ylab = "")
```

### Fit a logistic growth curve to the U.S. population data set

```{r}
pop.ss <- nls(population ~ SSlogis(year, phi1, phi2, phi3), data = USPop)
summary(pop.ss)
library(scales)

plot(population ~ year, USPop, xlim = c(1790, 2150),
     ylim = c(0, 500), las = 1, pch = "*",
     xlab = "Census year", ylab = "Population (millions)", cex = 1.6)
with(USPop, lines(seq(1790, 2150, by = 10),
                  predict(pop.ss, data.frame(year = seq(1790, 2150, by = 10))),
                  lwd = 1, col = alpha("blue", 0.75)))
abline(h = coef(pop.ss)[1], col = alpha("black", 0.7))
mtext(expression(hat(phi)[1]), side = 2, at = coef(pop.ss)[1], las = 1, col = "blue")
grid()
abline(v = coef(pop.ss)[2], col = alpha("black", 0.7), lwd = 0.5)
mtext(expression(hat(phi)[2]), side = 1, at = coef(pop.ss)[2], las = 1, col = "blue")
segments(coef(pop.ss)[2], 200, coef(pop.ss)[2] + 3 * coef(pop.ss)[3])
text(coef(pop.ss)[2], 200, "[", col = "blue")
text(coef(pop.ss)[2] + 3 * coef(pop.ss)[3], 200, "]", col = "blue")
text(coef(pop.ss)[2] + 1.5 * coef(pop.ss)[3], 180, expression(3*hat(phi)[3]), col = "blue")
# Compute AIC
AIC(pop.ss)
```

### Alternative model: fit quadratic/cubic polynomial regression

```{r}
pop.qm <- lm(population ~ poly(year, 2), USPop)
pop.cm <- lm(population ~ poly(year, 3), USPop)
summary(pop.cm)
## Model selection
AIC(pop.cm); AIC(pop.qm)
anova(pop.qm, pop.cm)
```

### Comparing the fits

```{r}
plot(population ~ year, USPop, xlim = c(1790, 2100),
     ylim = c(0, 500), las = 1, pch = "*", col = "blue",
     xlab = "Census year", ylab = "Population (millions)", cex = 1.6)
with(USPop, lines(seq(1790, 2100, by = 10),
                  predict(pop.ss, data.frame(year = seq(1790, 2100, by = 10))),
                  lwd = 1, col = alpha("black", 0.75)))
points(2010, 308, pch = "*", cex = 2, col = "red")
abline(h = coef(pop.ss)[1], lty = 3, col = "gray", lwd = 0.95)
with(USPop, lines(seq(1790, 2100, by = 10), predict(pop.cm, data.frame(year = seq(1790, 2100, by = 10))),
                  lwd = 1, lty = 2, col = alpha("black", 0.75)))
legend("bottomright", legend = c("NLR", "PolyR-3rd"), lty = c(1, 2), bty = "n")
```


