---
title: "DSA 8020 R Session 7: Logistic Regression and Poisson Regression"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 5.5
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Logistic Regression: Horseshoe Crab Malting


*Data Source:* Brockmann, H. J. (1996). Satellite male groups in horseshoe crabs, Limulus polyphemus. *Ethology*, 102(1), 1-21.

### Load the data 


```{r}
crab <- read.table("http://users.stat.ufl.edu/~aa/cda/data/Crabs.dat", header = T)
```

### Fit a Linear Regression 

Let's fit a simple linear regression using \texttt{weight} as the predictor

```{r}
lmFit <- lm(y ~ weight, data = crab)
par(mar = c(3.5, 3.5, 0.8, 0.6))
with(crab, plot(weight, y, pch = 16,
                cex = 0.75, las = 1, xlab = "", ylab = ""))
mtext("Weight (kg)", side = 1, line = 2)
abline(lmFit, col = "red")
```

### Fit a Logistic Regression

```{r}
logitFit <- glm(y ~ weight, data = crab, family = "binomial")
summary(logitFit)
# Plot the fits
rg <- range(crab$weight)
xg <- seq(rg[1], rg[2], 0.01)
pred <- predict(logitFit, newdata = data.frame(weight = xg), type = "response")
par(mar = c(3.5, 3.5, 0.8, 0.6))
plot(crab$weight, crab$y, pch = 16, cex = 0.75, las = 1, xlab = "", ylab = "")
mtext("Weight (kg)", side = 1, line = 2)
abline(lmFit, col = "red")
lines(xg, pred, col = "blue")
legend("bottomright", legend = c("SLR", "Logistic"),
       col = c("red", "blue"), lty = 1, bty = "n")
```

### Confidence Intervals

```{r}
# Normal approximation
est <- summary(logitFit)$coefficients
(CI_norm <- est[2, 1] + c(-1, 1) * qnorm(0.975) * est[2, 2])

# Profile likelihood CI
library(MASS)
(CI_prof <- confint(logitFit)[2,])

plot(1, type = "n", xlab = "", ylab = "",
     xlim = c(1, 2.6), ylim = c(-0.05, 0.1),
     yaxt = "n", main = expression(hat(beta[1])))
segments(CI_norm[1], 0, CI_norm[2], col = "red")
segments(CI_prof[1], 0.05, CI_prof[2], col = "blue")
abline(v = est[2, 1], lty = 2, col = "gray")
legend("topleft", legend = c("norm", "prof"),
       col = c("red", "blue"), lty = 1,
       lwd = 0.8, bty = "n")
```

### Prediction

```{r}
pred <- predict(logitFit, newdata = data.frame(weight = xg), se.fit = TRUE)

fit <- pred$fit; se <- pred$se.fit

plot(xg, exp(fit) / (1 + exp(fit)),
     type = "l", col = "blue", ylim = c(0, 1),
     las = 1, xlab = "Weight", ylab = expression(pi))
lines(xg, exp(fit + 1.96 * se) / (1 + exp(fit + 1.96 * se)), lty = 2)
lines(xg, exp(fit - 1.96 * se) / (1 + exp(fit - 1.96 * se)), lty = 2)
```

### Raw Residual plot

```{r}
res <- resid(logitFit, type = "response")
pred <- predict(logitFit)

plot(pred, res, col = "blue", pch = 16,
     xlab = "Predicted values",
     ylab = "Raw residuals")
abline(h = 0, lty = 2)
```


### Binned Residuals 

```{r,message=FALSE}
wt_bin <- findInterval(crab$weight, unique(quantile(crab$weight, seq(0, 1, length.out = 20 + 1))), rightmost.closed = TRUE)
library(dplyr)
crab.res <- mutate(crab, residuals = residuals(logitFit), Linpred = predict(logitFit), bin = wt_bin)

res_bin <- tapply(crab.res$residuals, crab.res$bin, mean)
Lpred_bin <- tapply(crab.res$Linpred, crab.res$bin, mean)

plot(res_bin ~ Lpred_bin, xlab = "Predicted values",
     ylab = "Binned residuals", col = "blue", pch = 16, las = 1)
abline(h = 0, lty = 2)
```

### Model selection

```{r}
logitFit2 <- glm(y ~ weight + width, data = crab, family = "binomial")
summary(logitFit2)
step(logitFit2)
```

## Poisson Regression

### Flying-Bomb Hits on London During World War II [Clarke, 1946; Feller, 1950]

```{r}
count <- c(229, 211, 93, 35, 7, 1)
grids <- 576
hits <- 537
lambda <- hits / grids
count_expected <- c(grids * dpois(0:4, lambda = lambda), grids * ppois(4, lambda = lambda, lower.tail = F))
round(count_expected, 1)
```

### US Landfalling Hurriances 

```{r}
# load the hurriance count
con = "http://myweb.fsu.edu/jelsner/Book/Chap07/US.txt"
hurricanes = read.table(con, header = T)
head(hurricanes)

par(las = 1, mar = c(4.6, 3.9, 0.8, 0.6))
layout(matrix(c(1, 2), 1, 2, byrow = TRUE), widths = c(0.57, 0.43))
plot(hurricanes$Year, hurricanes$All, type = "h", xlab = "Year", ylab = "Hurricane count")
grid()
barplot(table(hurricanes$All), xlab = "Hurricane Count", ylab = "Number of years", main = "")
```

### Load the environmetal variables

```{r}
load("annual.RData")
data <- data.frame(All = hurricanes$All, SOI = annual$soi,
                   NAO = annual$nao, SST = annual$sst,
                   SSN = annual$ssn)
data <- data[-(1:15),]
with(data, plot(All ~ SOI, las = 1))
```


```{r}
H <- hurricanes

par(mfrow = c(2, 2), mar = c(4.5, 4, 1, 0.6))
plot(range(annual$sst, na.rm = TRUE), c(0, 7), type = "n", ylab = "Hurricane count", xlab = "SST",
     las = 1)
for(i in 0:7){
  points(fivenum(annual$sst[H$All == i])[3], i, pch = 19)
  lines(c(fivenum(annual$sst[H$All == i])[1], fivenum(annual$sst[H$All == i])[2]), c(i, i))
  lines(c(fivenum(annual$sst[H$All == i])[4], fivenum(annual$sst[H$All == i])[5]), c(i, i))
}
plot(range(annual$soi, na.rm = TRUE), c(0, 7), type = "n", ylab = "Hurricane count", xlab = "SOI",
     las = 1)

for(i in 0:7){
  points(fivenum(annual$soi[H$All == i])[3], i, pch=19)
  lines(c(fivenum(annual$soi[H$All == i])[1], fivenum(annual$soi[H$All == i])[2]), c(i, i))
  lines(c(fivenum(annual$soi[H$All == i])[4], fivenum(annual$soi[H$All == i])[5]), c(i, i))
}
plot(range(annual$nao, na.rm = TRUE), c(0, 7), type = "n", ylab = "Hurricane count", xlab = "NAO",
     las = 1)

for(i in 0:7){
  points(fivenum(annual$nao[H$All == i])[3], i, pch=19)
  lines(c(fivenum(annual$nao[H$All == i])[1], fivenum(annual$nao[H$All == i])[2]), c(i, i))
  lines(c(fivenum(annual$nao[H$All == i])[4], fivenum(annual$nao[H$All == i])[5]), c(i, i))
}
plot(range(annual$ssn, na.rm = TRUE), c(0, 7), type = "n", ylab = "Hurricane count",
     xlab = "Sunspot number", las = 1)

for(i in 0:7){
  points(fivenum(annual$ssn[H$All == i])[3], i, pch = 19)
  lines(c(fivenum(annual$ssn[H$All == i])[1], fivenum(annual$ssn[H$All == i])[2]), c(i, i))
  lines(c(fivenum(annual$ssn[H$All == i])[4], fivenum(annual$ssn[H$All == i])[5]), c(i, i))
}
```

### Linear Regression 

```{r}
lmFull <- lm(All ~ ., data = data)
predict(lmFull, newdata = data.frame(SOI = -3, NAO = 3, SST = 0, SSN = 250))

plot(lmFull, which = 1)
plot(lmFull, which = 3)
```

### Poisson Regression

```{r}
PoiFull <- glm(All ~ ., data = data, family = "poisson")
summary(PoiFull)

plot(data$SOI, hurricanes$All[-(1:15)], cex = 0.75, col = "gray",
     xlab = "", ylab = "", las = 1)
mtext("Hurricane Count", side = 2, line = 2)
mtext("Year", side = 1, line = 2)
points(data$SOI, predict(lmFull), col = "red",
       cex = 0.5, pch = 16)
points(data$SOI, predict(PoiFull, type = "response"), col = "blue", cex = 0.5, pch = 16)
```

