---
title: "Extreme Value Analysis for Clemson Daily Rainfall"
author: "Whitney Huang"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

These data were taken from United States Historical Climatology Network（http://cdiac.ornl.gov/epubs/ndp/ushcn/ushcn.html). These data were compiled by Will Kleiber from CU-Boulder, thanks Will!

## Load the USHCN dataset

```{r}
load("USHCNprcpSetup.rdata")
```

## Extract the time series at a station near Clemson

```{r, message=FALSE}
Clemson.lon.lat <-  c(-82.8374, 34.6834)
library(fields)
dist2Clemson <- rdist.earth(matrix(Clemson.lon.lat, 1, 2), lon.lat, miles = F)
id <- which.min(dist2Clemson)
lon.lat[id,] # near Lamaster Dairy Center
stn.name[id]

## Use data from 1950/1/1-2014/12/31
dat <- PRCP[(50 * 365 + 1):41975, id]
```

```{r}
# Reformat the data into [day, year] and convert rainfall to inches
rain <- array(dat, dim = c(365, 65)) / 100
# Extract annual maxima
annmx <- apply(rain, 2, max, na.rm = T)
# Extract the timing of annual maxima
annmxT <- apply(rain, 2, which.max)
# Plot the daily time series
plot(1950 + 1:23725/365, c(rain),
     xlab = "Year", ylab = "Precipitation (in)",
     main = "Daily Precip in Clemson", type = "h",
     pch = 19, cex = 0.5, col = "lightblue",
     ylim = c(0, 7), las = 1)
grid()
```

## Block-Maxima Method 

We are going to conduct an extreme value analysis using the `extRemes` package developed and maintained by `Eric Gilleland`.

### Step I: Determine the block size and compute maxima for blocks

```{r, message=FALSE}
# Setting up the figure configuration
old.par <- par(no.readonly = TRUE)
mar.default <- par("mar")
mar.left <- mar.default
mar.right <- mar.default
mar.left[2] <- 0
mar.right[4] <- 0

# Time series plot
par(fig = c(0.2, 1, 0, 1), mar = mar.left)
plot(1950 + 1:23725/365, c(rain),
      xlab = "Year", ylab = "",
      main = "Daily Precip in Clemson",
      type = "h", pch = 19, cex = 0.5, col = "lightblue",
      ylim = c(0, 7.5), yaxt = "n")
par(las = 2)
axis(4, at = 0:8)
par(las = 0)
mtext("Precipitation (in)", side = 2, line = 5)
abline(v = 1950:2015, col = "gray", lty = 2)
points(1950:2014 + annmxT / 365, annmx,
       pch = 16, col = "blue", cex = 0.5)
# Histogram
hs <- hist(annmx,
           breaks = seq(0, 7.5, length.out = 35),
           plot = FALSE)

par(fig = c(0, 0.2, 0, 1.0), mar = mar.right, new = T)
plot (NA, type = 'n', axes = FALSE, yaxt = 'n',
      col = rgb(0, 0, 0.5, alpha = 0.5),
      xlab = "Density", ylab = NA, main = NA,
      xlim = c(-0.55, 0),
      ylim = c(0, 7.5))
axis(1, at = c(-0.5, -0.25, 0), c(0.5, 0.25, 0), las = 2)
arrows(rep(0, length(hs$breaks[-35])), hs$breaks[-35],
       -hs$density, hs$breaks[-35], col = "blue",
       length = 0, angle = 0, lwd = 1)
arrows(rep(0, length(hs$breaks[-1])), hs$breaks[-1],
       -hs$density, hs$breaks[-1], col = "blue",
       length = 0, angle = 0, lwd = 1)
arrows(-hs$density, hs$breaks[-35], -hs$density,
       hs$breaks[-1], col = "blue", angle = 0,
       length = 0)
library(extRemes)
mle <- fevd(annmx)$results$par
xg <- seq(0, 7.5, length.out = 100)
library(ismev)
lines(-gev.dens(mle, xg), xg, col = "red") 
par(old.par)
```

### Step II: Fit a GEV to the maxima and assess the fit

We assume $m_{1}, \cdots, m_{t}$ follows a GEV distribution $GEV(\mu_{n}, \sigma_{n}, \xi)$ and we maximize the log-likelihood  
$$-\sum_{i=1}^{t} m_{i}^{-\frac{1}{\xi}}- t log(\sigma) - (\frac{1}{\xi}+1)\sum_{i=1}^{t}log(m_{i})$$


```{r, message=FALSE, fig.height=6, fig.width=6}
# Fit a GEV to annual maximum daily precip using MLE
library(scales)
gevfit <- fevd(annmx)
# Print the results 
gevfit
#QQ plot
p <- 1:65 / 66
qm <- gevq(mle, 1 - p)
plot(qm, sort(annmx), xlim = c(1.8, 6.5), ylim = c(1.8, 6.5), pch = 16, cex = 1, col = alpha("blue", 0.5), xlab = "Model", ylab = "Empirical",
     main ="Quantile Plot", las = 1)
abline(0, 1, lwd = 1.5)
```

### Step III: Perform inference for return levels

Suppose we are interested in estimating 50-year return level



```{r}
RL50 <- return.level(gevfit, return.period = 50) # Estimate of the 100-year event
RL50
# Quantify the estimate uncertainty
## Delta method
CI_delta <- ci(gevfit, return.period = 50, verbose = T) 
CI_delta
## Profile likelihood method
CI_prof <- ci(gevfit, method = "proflik", xrange = c(4, 10),
   return.period = 50, verbose = F)
CI_prof 

hist(annmx, breaks = seq(0, 7.5, length.out = 35),
     col = alpha("lightblue", 0.2), border = "gray",
     xlim = c(0, 7.5), prob = T, ylim = c(0, 0.6),
     xlab = " Annmx Precip (in)",
     main = "95% CI for 50-yr RL",
     las = 1)
xg <- seq(0, 10, len = 1000)
mle <- gevfit$results$par
lines(xg, gev.dens(mle, xg), lwd = 1.5)
for (i in c(1, 3)) abline(v = CI_delta[i], lty = 2, col = "blue")
for (i in c(1, 3)) abline(v = CI_prof[i], lty = 2, col = "red")
abline(v = RL50, lwd = 1.5, lty = 2)
legend("topleft", legend = c("Delta CI", "Prof CI"), col = c("blue", "red"),
       lty = c(2, 3), bty = "n", cex = 1.25, lwd = 1.5)
```


## Peak-Over-Threshold Method


### Step I: Pick a threshold and extract the threshold exceedances


```{r}
old.par <- par(no.readonly = TRUE)
mar.default <- par('mar')
mar.left <- mar.default
mar.right <- mar.default
mar.left[2] <- 0
mar.right[4] <- 0
# Time series plot
par(fig = c(0.2, 1, 0, 1), mar = mar.left)
plot(1950 + 1:23725/365, c(rain),
      xlab = "Year", ylab = "",
      main = "Daily Precip in Clemson",
      type = "h", pch = 19, cex = 0.5, col = "lightblue",
      ylim = c(0, 7.5), yaxt = "n")
par(las = 2)
axis(4, at = 0:8)
par(las = 0)
mtext("Precipitation (in)", side = 2, line = 4)
#Threshold exceedances
thres <- 0.75
ex_id <- which(rain > thres)
ex <- rain[ex_id]
length(ex)
#Extract the timing of POT
abline(h = thres, col = "blue", lty = 2)
points(1950 + ex_id / 365, ex, col = alpha("blue", 0.5), pch = 16,
       cex = 0.75)
par(las = 2)
axis(4, at = 0:8)
par(las = 0)
#mtext("Precipitation (in)", side = 2, line = 5)
grid()
hs <- hist(ex, seq(thres, 7.5, len = 50), plot = FALSE)
par(fig = c(0, 0.2, 0, 1.0), mar = mar.right, new = T)
plot (NA, type = 'n', axes = FALSE, yaxt = 'n',
      col = rgb(0,0,0.5, alpha = 0.5),
      xlab = "Density", ylab = NA, main = NA,
      xlim = c(-max(hs$density), 0),
      ylim = c(0, 7.5))
axis(1, at = c(-1.6, -1.2, -0.8, -0.4, 0),
     c(1.6, 1.2, 0.8, 0.4, 0), las = 2)
#abline(h = 21, col = "red", lty = 5)
arrows(rep(0, length(hs$breaks[-50])), hs$breaks[-50],
       -hs$density, hs$breaks[-50], col = "blue",
       length = 0, angle = 0, lwd = 1)
arrows(rep(0, length(hs$breaks[-1])), hs$breaks[-1],
       -hs$density, hs$breaks[-1], col = "blue",
       length = 0, angle = 0, lwd = 1)
arrows(-hs$density, hs$breaks[-50], -hs$density,
       hs$breaks[-1], col = "blue", angle = 0,
       length = 0)

mle <- fevd(c(rain[!is.na(rain)]), threshold = thres, type = "GP")$results$par
xg <- seq(thres, 7.5, length.out = 100)
lines(-gpd.dens(mle, thres, xg), xg, col = "red")
par(old.par)
```

### How to choose the "right" threshold?


```{r}
mrlplot(dat[!is.na(dat)] / 100, main = "Mean Residual Life", xlab = "Threshold (in)", las = 1)
# I choose 0.75 as the threshold but note that the "straightness"
# is difficult to assess
abline(v = 0.75, col = "blue", lty = 2)
```


### Step II: Fit a GPD to threshold excesses and assess the fit

```{r, fig.height=6, fig.width=6}
# Fit a GPD for threshold exceenances using MLE
gpdfit1 <- fevd(c(rain[!is.na(rain)]), threshold = thres, type = "GP")
# Print the results 
gpdfit1
# QQ plot
p <- 1:1489 / 1490
qm <- gpdq(mle, 0.75, 1 - p)
plot(qm, sort(ex), xlim = c(0.5, 6.5), ylim = c(0.5, 6.5),
     pch = 16, cex = 1, col = alpha("blue", 0.5),
     xlab = "Model", ylab = "Empirical", main = "Quantile Plot",
     las = 1)
abline(0, 1, lwd = 1.5)
```


### Step III: Perform inference for return levels

Again we are interested in estimating 50-year return level


```{r}
RL50 <- return.level(gpdfit1, return.period = 50 * 365 / 365.25)
RL50
CI_delta <- ci(gpdfit1, return.period = 50 * 365 / 365.25,
               verbose = F) 
CI_delta
CI_prof <- ci(gpdfit1, method = "proflik", xrange = c(4, 10),
   return.period = 50 * 365 / 365.25, verbose = F)
CI_prof


hist(ex, 40, col = alpha("lightblue", 0.2), border = "gray",
     xlim = c(thres, 7.5), prob = T, ylim = c(0, 2),
     xlab = "Threshold excess (in)",
     main = "95% CI for 50-yr RL")
xg <- seq(thres, 10, len = 1000)
mle <- gpdfit1$results$par
lines(xg, gpd.dens(mle, thres, xg), lwd = 1.5)
for (i in c(1, 3)) abline(v = CI_delta[i], lty = 2, col = "blue")
for (i in c(1,3)) abline(v = CI_prof[i], lty = 2, col = "red")
abline(v = RL50, lwd = 1.5, lty = 2)
legend("topleft", legend = c("Delta CI", "Prof CI"),
       col = c("blue", "red"), lty = c(2, 3), bty = "n", cex = 1.25,
       lwd = 1.5)
```
