---
title: "Multivariate Extremes"
author: "Whitney Huang"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Motiviating (simulated) example 

```{r}
library(MASS)
# Simulate bivariate normal with corr = 0.75
rho = 0.75
set.seed(123)
sim <- mvrnorm(n = 1000, mu = rep(0, 2), 
               Sigma = matrix(c(1, rho, rho, 1), 2, 2))
cor(sim)
# Load R package evd
library(evd)
# Sample size
n <- 1000
set.seed(123)
# Simulate bivariate extreme random vector with 
# unit Frechet and logistic copula (alpha = 0.5)
x <- rmvevd(n, dep = .5, model = "log", mar = c(0, 1, 0))
cor(x)
# transfer the marginals to standard normal
u1 <- pgev(x[, 1]); n1 <- qnorm(u1)
u2 <- pgev(x[, 2]); n2 <- qnorm(u2)
cor(cbind(n1, n2))
```

## Plot the data 

```{r}
par(mfrow = c(1, 2), las = 1, mgp = c(2, 1, 0),
    mar = c(3.6, 3.6, 0.8, 0.6))
plot(sim, pch = 16, cex = 0.5,
     xlab = expression(X[1]),
     ylab = expression(X[2]), xlim = c(-3.75, 3.75),
     ylim = c(-3.75, 3.75))
grid()
legend("topleft", legend = round(cor(sim)[1, 2], 3),
       title = expression(hat(rho)), text.col = "blue",
       bty = "n")
plot(cbind(n1, n2), pch = 16, cex = 0.5,
     xlab = expression(Y[1]),
     ylab = expression(Y[2]), xlim = c(-3.75, 3.75),
     ylim = c(-3.75, 3.75))
grid()
legend("topleft", legend = round(cor(n1, n2), 3),
       title = expression(hat(rho)), text.col = "blue",
       bty = "n")
```

## Multivariate extreme is not uniquely defined

```{r}
par(mfrow = c(2, 2), las = 1, mgp = c(2, 1, 0),
    mar = c(3.6, 3.6, 1, 0.6))
plot(sim, pch = 16, cex = 0.5,
     xlab = expression(X[1]),
     ylab = expression(X[2]), xlim = c(-3.75, 3.75),
     ylim = c(-3.75, 3.75))
points(max(sim[, 1]), max(sim[, 2]), pch = "*",
       cex = 2, col = "red")
abline(h = max(sim[, 2]), lty = 2, col = "red")
abline(v = max(sim[, 1]), lty = 2, col = "red")
grid()
mtext("Componentwise maxima", col = "red")
plot(sim, pch = 16, cex = 0.5,
     xlab = expression(X[1]),
     ylab = expression(X[2]), xlim = c(-3.75, 3.75),
     ylim = c(-3.75, 3.75))
ex1 <- which(sim[, 1] >= 2.8)
ex2 <- which(sim[, 2] >= 2.8)
ex <- intersect(ex1, ex2)
points(sim[ex, 1], sim[ex, 2], pch = "*", cex = 2, col = "red")
abline(h = 2.8, lty = 2, col = "gray")
abline(v = 2.8, lty = 2, col = "gray")
grid()
mtext("Marginal thresholds", col = "red")
plot(sim, pch = 16, cex = 0.5,
     xlab = expression(X[1]),
     ylab = expression(X[2]), xlim = c(-3.75, 3.75),
     ylim = c(-3.75, 3.75))
points(sim[ex1, 1], sim[ex1, 2], pch = "*", cex = 2, col = "blue")
points(sim[ex2, 1], sim[ex2, 2], pch = "*", cex = 2, col = "green")
points(sim[ex, 1], sim[ex, 2], pch = "*", cex = 2, col = "red")
abline(h = 2.8, lty = 2, col = "gray")
abline(v = 2.8, lty = 2, col = "gray")
grid()
mtext("Concomitants", col = "red")
plot(sim, pch = 16, cex = 0.5,
     xlab = expression(X[1]),
     ylab = expression(X[2]), xlim = c(-3.75, 3.75),
     ylim = c(-3.75, 3.75))
l1 <- which(sim[, 1] + sim[, 2] >=4.5)
points(sim[l1, 1], sim[l1, 2], pch = "*", cex = 2, col = "red")
abline(a = 4.5, b = -1, col = "red")
grid()
mtext("Norm Threshold (L1)", col = "red")
```

## Plot $\chi(u)$

```{r}
par(mfrow = c(1, 2), las = 1, mgp = c(2, 1, 0),
    mar = c(3.6, 3.6, 1, 0.6))
chiplot(sim, which = 1, ylim1 = c(0, 1), nq = 60, col = "red")
chiplot(cbind(n1, n2), which = 1, ylim1 = c(0, 1), nq = 60, col = "red")
```

## Annual Maximum Wind Speeds at Albany and Hartford

```{r}
library(ismev)
data(wind); head(wind)
hartford <- wind[, 2]; albany <- wind[, 3]
blockmax <- cbind(hartford, albany)
```

## Plot the componentwise maxima "data"

```{r}
par(las = 1, mgp = c(2, 1, 0), mar = c(3.6, 3.6, 1, 0.6))
plot(blockmax, pch = 16, cex = 0.65,
     xlab = "Ann Max Wind Speed at Albany (Knot)",
     ylab = "Ann Max Wind Speed at Hartford (Knot)")
grid()
```

## Fit bivariate GEV

```{r}
fit1 <- fbvevd(blockmax, model = "log")
fit2 <- fbvevd(blockmax, model = "bilog")

par(las = 1, mgp = c(2, 1, 0), mar = c(3.6, 3.6, 1, 0.6), mfrow = c(2, 2))
plot(fit1, which = 6,
     xlab = expression(omega), ylab = expression(h(omega)))
plot(fit2, which = 6,
     xlab = expression(omega), ylab = expression(h(omega)))
plot(fit1, which = 3, pch = 16, cex = 0.35, xlab = "Hartford", ylab = "Albany")
plot(fit2, which = 3, pch = 16, cex = 0.35, xlab = "Hartford", ylab = "Albany")
```

## Wave and Surge Heights in South-West England

Let's plot the data as well as the $\chi(u)$ plot

```{r}
data(wavesurge); head(wavesurge)
wave <- wavesurge[, 1]; surge <- wavesurge[, 2]

par(las = 1, mgp = c(2.2, 1, 0), mar = c(3.6, 3.6, 0.8, 0.6))
plot(wave, surge, pch = 16, cex = 0.5,
     xlab = "Wave (m)", ylab = "Surge (m)")
grid()

chiplot(wavesurge)
```

## Fit bivariate threshold excess models

```{r}
thresh <- c(quantile(wave, 0.95), quantile(surge, 0.95))
par(las = 1, mgp = c(2.2, 1, 0), mar = c(3.6, 3.6, 0.8, 0.6))
plot(wave, surge, pch = 16, cex = 0.5,
     xlab = "Wave (m)", ylab = "Surge (m)")
grid()
abline(v = thresh[1], col = "red")
abline(h = thresh[2], col = "red")

par(las = 1, mgp = c(2.2, 1, 0), mar = c(3.6, 3.6, 0.8, 0.6))
plot(wave, surge, pch = 16, cex = 0.5,
     xlab = "Wave (m)", ylab = "Surge (m)")
grid()
abline(v = thresh[1], col = "red")
abline(h = thresh[2], col = "red")
legend("topleft", legend = "Region 4", bty = "n", text.col = "blue")
legend("topright", legend = "Region 1", bty = "n", text.col = "blue")
legend("bottomleft", legend = "Region 3", bty = "n", text.col = "blue")
legend("bottomright", legend = "Region 2", bty = "n", text.col = "blue")

fitlogistic <- fbvpot(wavesurge, thresh)
fitbilogistic <- fbvpot(wavesurge, thresh, model = "bilog")

par(las = 1, mgp = c(2.4, 1, 0), mar = c(3.6, 3.6, 1, 0.6), mfrow = c(2, 2))
plot(fitlogistic, which = 4,
     xlab = expression(omega), ylab = expression(h(omega)))
plot(fitbilogistic, which = 4,
     xlab = expression(omega), ylab = expression(h(omega)))
plot(fitlogistic, which = 1, pch = "*", xlab = "Wave (m)", ylab = "Surge (m)",
     col = "red")
plot(fitbilogistic, which = 1, pch = "*", xlab = "Wave (m)", ylab = "Surge (m)", col = "red")
```


## Conditional extreme values modelling [Heffernan and Tawn, 2004]

Air pollution data from Leeds (U.K.) city centre, collected from 1994 to 1998. The summer data set corresponds to the months of April to July inclusive. The winter data set corresponds to the months of November to February inclusive. Some outliers have been removed, as discussed by Heffernan and Tawn, 2004. Here we only consider $\mathrm{NO}_{2}$ and $\mathrm{O}_{3}$ during summer to illustrate the CEV modeling framework. 


```{r,message=FALSE}
library(texmex)
data(winter); data(summer)
(CEV_Fit <- mex(summer[, 1:2], mqu = .9, dqu = 0.85, which = "O3"))
par(las = 1, mgp = c(2.4, 1, 0), mar = c(3.6, 3.6, 1, 0.6), mfrow = c(1, 3))
plot(CEV_Fit)
```




