---
title: "Spatial Extremes"
author: "Whitney Huang"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

**Acknowledgment:** This R session is adopted from Ribatet's short course on "Modelling spatial extremes with the SpatialExtremes package".

## Spatial GEV

### Plot the spatial locations

```{r}
library(SpatialExtremes)
data(wind)
par(mar = rep(0, 4))
maps::map(xlim = c(0, 9), ylim = c(47.5, 56))
points(coord, pch = 15, cex = 0.5)
```

### Fitting spatial GEV

```{r}
loc.form <- y ~ lon * lat; scale.form <- shape.form <- y ~ 1
#Convert coordinates from degrees to km (rough)
coord[, 1:2] <- 111 * coord[, 1:2]
(M0 <- fitspatgev(wind, scale(coord, scale=FALSE), loc.form, scale.form, shape.form))
```

### "Prediction"

```{r,message=FALSE}
library(fields)
x <- seq(min(coord[,1]), max(coord[,1]), length = 100)
y <- seq(min(coord[,2]), max(coord[,2]), length = 100)
grid <- expand.grid(x, y); colnames(grid) <- c("lon", "lat")
## Watch out we scale the data so we need to use the same transformation 
grid[,1] <- grid[, 1] - mean(coord[, 1])
grid[,2] <- grid[, 2] - mean(coord[, 2])
ans <- predict(M0, newdata = grid, ret.per = 20)$Q20
## Do the plot
maps::map(xlim = range(x) / 111, ylim = range(y) / 111)
image(x / 111, y / 111, matrix(ans, 100), add = TRUE, col = tim.colors(64))
contour(x / 111, y / 111, matrix(ans, 100), add = TRUE); maps::map(add = TRUE)
```

## Bayesian Hierarchical Model

### Plot the spatial domain and station locations
```{r}
data(rainfall)
data(swissalt)
par(mar = rep(0, 4), ps = 16)
image(lon.vec, lat.vec, alt.mat, col = terrain.colors(64), asp = 1,
      bty = "n", xlab = "n", ylab = "n", axes = FALSE)
swiss(add = TRUE, city = TRUE)
points(coord, pch = 16, cex = 0.75, col = "blue")
```

### Prior specification

```{r}
hyper <- list()
hyper$betaMeans <- list(loc = rep(0, 3), scale = rep(0, 3), shape = 0)
hyper$betaIcov <- list(loc = diag(rep(1/1000, 3)), scale = diag(rep(1 / 1000, 3)), shape = 1 / 10)
hyper$sills <- list(loc = c(1, 12), scale = c(1, 1), shape = c(1, 0.04))
hyper$ranges <- list(loc = c(5, 3), scale = c(5, 3), shape = c(5, 3))
hyper$smooths <- list(loc = c(1, 1), scale = c(1, 1), shape = c(1, 1))
prop <- list(gev = c(3, 0.1, 0.3), ranges = c(1, 0.8, 1.2), smooths = rep(0, 3))
start <- list(sills = c(10, 10, 0.5), ranges = c(20, 10, 10), smooths = c(1, 1, 1), beta = list(loc = c(25, 0, 0), scale = c(33, 0, 0), shape = 0.001))
```

### Running the Gibbs sampler

```{r}
loc.form <- scale.form <- y ~ lon + lat; shape.form <- y ~ 1
chain <- latent(rain, coord[, 1:2], "powexp", loc.form, scale.form, shape.form,
                hyper = hyper, prop = prop,
start = start, n = 100, burn.in = 500, thin = 5)
chain
```

### Making 20-year return level map

```{r}
map.latent(chain, ret.per = 20, plot.contour = FALSE, col = tim.colors())
```


## Max-stable processes

### USHCN data

```{r}
data(USHCNTemp)
maps::map("usa")
points(metadata[, c("lon","lat")], pch = 16, cex = 0.75)
```

### F– madogram

```{r}
data <- maxima.summer
coord <- as.matrix(metadata[, c("lon", "lat")])
fmadogram(data, coord)
fmadogram(data, coord, which = 'ext', n.bins = 100, pch = 16)
abline(h = 1, lty = 2); abline(h = 2, lty = 2)
```


### Simulation from various max-stable models

```{r}
x <- seq(0, 10, length = 100)
n.sim <- 5
schlather <- rmaxstab(n.sim, x, "powexp", nugget = 0, range = 3, smooth = 1)
extremalt <- rmaxstab(n.sim, x, "tpowexp", DoF = 4, nugget = 0, range = 3, smooth = 1)
brown <- rmaxstab(n.sim, x, "brown", range = 3, smooth = 1)
smith <- rmaxstab(n.sim, x, "gauss", var = 2)

plot(x, schlather[1,], type = "l", ylim = range(schlather))
for (i in 2:5) lines(x, schlather[i,], col = i)
plot(x, smith[1,], type = "l", ylim = range(smith))
for (i in 2:5) lines(x, smith[i,], col = i)
```

### Two-step max-stable process fitting

```{r}
data(rainfall)
alt <- coord[,3]; coord <- coord[,-3]
## Transformation to unit Frechet margins
rain.frech <- apply(rain, 2, gev2frech, emp = TRUE)
(fit <- fitmaxstab(rain.frech, coord, "powexp", nugget = 0))
```

### One-step model fitting

```{r}
loc.form <- y ~ lon + lat + alt
scale.form <- y ~ lon + lat
shape.form <- y ~ 1
(fit <- fitmaxstab(rain, coord, "powexp", nugget=0, loc.form, scale.form, shape.form,
marg.cov = cbind(alt = alt)))
plot(fit)
```

### LRT Model selection 

```{r}
M0 <- fitmaxstab(rain, coord, "powexp", nugget = 0, locCoeff4 = 0, loc.form, scale.form,
                 shape.form, marg.cov = cbind(alt = alt))
M1 <- fitmaxstab(rain, coord, "powexp", nugget = 0,
loc.form, scale.form, shape.form, marg.cov = cbind(alt = alt))
anova(M0, M1)
```

### Model selection via TIC

```{r}
M1 <- fitmaxstab(rain, coord, "powexp", nugget = 0, loc.form, scale.form, shape.form,
                 marg.cov = cbind(alt = alt))
M2 <- fitmaxstab(rain, coord, "whitmat", loc.form, scale.form,
                 shape.form, marg.cov = cbind(alt = alt))
TIC(M1, M2)
```

### Parameters and return level maps

```{r}
xlim <- range(M2$coord[,1]); ylim <- range(M2$coord[,2])
data(swissalt)##Retrieve altitude at other places
idx.x <- which(lon.vec >= xlim[1] & lon.vec <= xlim[2])
idx.y <- which(lat.vec >= ylim[1] & lat.vec <= ylim[2])
x <- lon.vec[idx.x]; y <- lat.vec[idx.y]
n.x <- length(x); n.y <- length(y)
covariates <- array(alt.mat[idx.x, idx.y], dim = c(n.x, n.y, 1), dimnames = list(NULL, NULL, "alt"))
covariates[is.na(covariates)] <- 0
par(mar = rep(0, 4), mfrow = c(1, 3))
SpatialExtremes::map(M2, x, y, covariates, param = "loc",
                     xaxt = "n", yaxt = "n", col = tim.colors()); swiss(add = TRUE)
SpatialExtremes::map(M2, x, y, covariates, param = "scale",
                     xaxt = "n", yaxt = "n", col = tim.colors()); swiss(add = TRUE)
SpatialExtremes::map(M2, x, y, covariates, param = "quant", ret.per = 20,
                     xaxt = "n", yaxt = "n", col = tim.colors()); swiss(add = TRUE)
```



