---
title: 'STAT 8010 R Lab 1: Simple Linear Regression'
author: "Whitney Huang"
date: "8/20/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab Objective

* To gain experience with R, a programming language and free software environment for statistical computing and graphics.
* To learn how to use R to conduct a simple linear regression analysis

### Example: Maximum Heart Rate vs. Age

The maximum heart rate ($\text{HR}_{max}$) of a person is often said to be related to age (Age) by the equation:
$$\text{HR}_{max} = 220 - \text{Age}$$

Let's use a dataset to assess this statement.


### Setup

* You should have R installed, if not, open a web browser and go to (http://cran.r-project.org) and download and install R. It also helpful to install RStudo (http://rstudio.com).

* Create a folder for this R lab. Download the Maximum Heart Rate dataset at (http://whitneyhuang83.github.io/maxHeartRate.csv) and save it in the folder you just created.

### Load the dataset

There are several ways to load a dataset into R:

+ Importing Data over the Internet

```{r}
dat <-read.csv('http://whitneyhuang83.github.io/STAT8010/Data/maxHeartRate.csv', header = T) 
```

Let's take a look at the data

```{r}
dat
```

+ Read the dataset from you computer

```{r, message=FALSE}
dat <- read.csv('maxHeartRate.csv', header = T) 
```

+ If the data is not too big, you can type the data into R

```{r}
age <- c(18, 23, 25, 35, 65, 54, 34, 56, 72, 19, 23, 42, 18, 39, 37)
maxHeartRate <- c(202, 186, 187, 180, 156, 169, 174, 172, 153,
                  199, 193, 174, 198, 183, 178)
dat <- data.frame(cbind(age, maxHeartRate))
```

### Examine the data before fitting models

```{r}
summary(dat)
var(dat$age); var(dat$maxHeartRate)
cov(dat$age, dat$maxHeartRate)
cor(dat$age, dat$maxHeartRate)
```

### Plot the data before fitting models

This is what the scatterplot would look like by default. Put predictor (age) to the first argument and response (maxHeartRate) to the second argument.


```{r}
plot(dat$age, dat$maxHeartRate)
```

Let's make the plot look nicer (type ?plot to learn more).

```{r}
par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(dat$age, dat$maxHeartRate,
     pch = 16, xlab = "Age",
     ylab = "Max heart rate (bpm)")
grid()
```

**Question:** Describe the direction, strength, and the form of the relationship.

### Simple linear regression

Let's do the calculations to figure out the regression coefficients as well as the standard deviation of the random error.   

* Slope

```{r}
X <- dat$age; Y <- dat$maxHeartRate
Y_diff <- Y - mean(Y)
X_diff <- X - mean(X)
beta_1 <- sum(Y_diff * X_diff) / sum((X_diff)^2)
beta_1
```
* Intercept

```{r}
beta_0 <- mean(Y) - mean(X) * beta_1
beta_0
```


* Fitted values
```{r}
Y_hat <- beta_0 + beta_1 * X
Y_hat
```

* $\hat{\sigma}$

```{r}
sigma2 <- sum((Y - Y_hat)^2) / (length(Y) - 2)
sqrt(sigma2)
```

Add the fitted regression line to the scatterplot

```{r}
par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(dat$age, dat$maxHeartRate,
     pch = 16, xlab = "Age",
     ylab = "Max heart rate (bpm)")
grid()
abline(a = beta_0, b = beta_1,
       col = "blue")
```

### Let R do all the work

```{r}
fit <- lm(maxHeartRate ~ age,
          data = dat)
summary(fit)
```

* Regression coefficients

```{r}
fit$coefficients
```
* Fitted values

```{r}
fit$fitted.values
```

* $\hat{\sigma}$

```{r}
summary(fit)$sigma
```





