---
title: "STAT 8020 R Lab 3: Simple Linear Regression III"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Understanding Sampling Distributions and Confident Intervals via simulation

Simulate the "data" $\{x_{i}, y_{i}\}_{i=1}^{n}$ where $y_{i} = \beta_{0}+\beta_{1}x_{i} + \varepsilon_{i}$, $\varepsilon \sim \mathrm{N}(0, \sigma^2)$. Repeat this process $N$ times.

### Generate data in R

```{r}
set.seed(12)
n = 30; beta0 = 3; beta1 = 1.5; N = 100; sigma2 = 1 
x <- 3 + 5 * runif(n)
set.seed(123)
y <- replicate(N, beta0 + beta1 * x + rnorm(n, mean = 0, sd = sqrt(sigma2)))
dim(y)
```

### Plot the first few simulated datasets

```{r}
for (i in 1:5){
  plot(x, y[, i], pch = 16, las = 1, ylab = "y")
  abline(3, 1.5, col = "blue", lwd = 1.5)
}
```

### Estimate the $\beta_{0}$, $\beta_{1}$, and $\sigma^2$ for each simulated dataset 

```{r}
beta0_hat <- beta1_hat <- sigma2_hat <- se_beta1 <- numeric(N)
for (i in 1:100){
  fit <- lm(lm(y[, i] ~ x))
  beta0_hat[i] <- summary(fit)[["coefficients"]][, 1][1]
  beta1_hat[i] <- summary(fit)[["coefficients"]][, 1][2]
  se_beta1[i] <- summary(fit)[["coefficients"]][, 2][2]
  sigma2_hat[i] <- summary(fit)[["sigma"]]^2
}
```

### Assess the estimation perfromance 

```{r}
boxplot(beta0_hat, las = 1, main = expression(hat(beta[0])))
abline(h = beta0, col = "blue", lwd = 1.5)

boxplot(beta1_hat, las = 1, main = expression(hat(beta[1])))
abline(h = beta1, col = "blue", lwd = 1.5)

boxplot(sigma2_hat, las = 1, main = expression(paste("Boxplot of ", hat(sigma)^2)))
abline(h = sigma2, col = "blue", lwd = 1.5)
```

### Sampling distribution

```{r}
hist(beta1_hat, 16, col = "lightblue", border = "gray",
     main = expression(paste("Histogram of ", hat(beta)[1])),
     xlab = expression(hat(beta)[1]))
abline(v = beta1, col = "blue", lwd = 1.5)
mtext(expression(beta[1]), 1, at = beta1, col = "blue")
```

### CI's for all the simulated datasets

```{r}
t <- qt(1 - 0.05 / 2, n - 2)
LL <- beta1_hat - t * se_beta1
UL <- beta1_hat + t * se_beta1
miss <- which((LL - beta1) * (UL - beta1) > 0)


par(las = 1)
plot(1:100, rep(beta1, N), type = "l", bty = "n", xaxt = "n", xlab = "",
     lwd = 1.5, ylab = expression(hat(beta)[1]))
for (i in 1:100){
  segments(i, LL[i], i, UL[i], col = "blue")
}

for (i in miss){
  segments(i, LL[i], i, UL[i], col = "red")
}
```


## Maximum Heart Rate vs. Age Example

### First Step: Load the data

```{r}
dat <- read.csv('http://whitneyhuang83.github.io/STAT8010/Data/maxHeartRate.csv', header = T) 
head(dat)
attach(dat)
```

## Fitting a simple linear regression

```{r}
fit <- lm(MaxHeartRate ~ Age)
summary(fit)
```


## Confidence Interval
### $\beta_{1}$

```{r}
alpha = 0.05
beta1_hat <- summary(fit)[["coefficients"]][, 1][2]
se_beta1 <- summary(fit)[["coefficients"]][, 2][2]
CI_beta1 <- c(beta1_hat - qt(1 - alpha / 2, 13) * se_beta1,
              beta1_hat + qt(1 - alpha / 2, 13) * se_beta1)
CI_beta1
```

### $Y_{h}|X_{h} = 40$

```{r}
Age_new = data.frame(Age = 40)
hat_Y <- fit$coefficients[1] + fit$coefficients[2] * 40
hat_Y
predict(fit, Age_new, interval = "confidence") 
predict(fit, Age_new, interval = "predict")
```


### Check

```{r}
sd <- sqrt((sum(fit$residuals^2) / 13))
ME <- qt(1 - alpha / 2, 13) * sd * sqrt(1 + 1 / 15 + (40 - mean(Age))^(2) / sum((Age - mean(Age))^2))
c(hat_Y - ME, hat_Y + ME)
```


```{r}
Age_grid = data.frame(Age = 18:72)
CI_band <- predict(fit, Age_grid, interval = "confidence") 
PI_band <- predict(fit, Age_grid, interval = "predict")

plot(dat$Age, dat$MaxHeartRate, pch = 16, cex = 0.75,
     xlab = "Age (Yr)", ylab = "Maximum Heart Rate (bpm)", las = 1)
abline(fit, col = "blue")
abline(v = mean(dat$Age), lty = 2, col = "gray")
abline(h = mean(dat$MaxHeartRate), lty = 2, col = "gray")
lines(18:72, CI_band[, 2], lty = 2, col = "blue")
lines(18:72, CI_band[, 3], lty = 2, col = "blue")
lines(18:72, PI_band[, 2], lty = 2, col = "darkgreen")
lines(18:72, PI_band[, 3], lty = 2, col = "darkgreen")
legend("topright", legend = c("CI", "PI"), col = c("blue", "darkgreen"),
       lty = 2, bty = "n")
```




## Hypothesis Tests for $\beta_{1}$

$H_{0}: \beta_{1} = -1$ vs. $H_{a}: \beta_{1} \neq -1$ with $\alpha = 0.05$

```{r}
beta1_null <- -1
t_star <- (beta1_hat - beta1_null) / se_beta1
p_value <- 2 * pt(t_star, 13, lower.tail = F) 
p_value


par(las = 1)
x_grid <- seq(-3.75, 3.75, 0.01)
y_grid <- dt(x_grid, 13)
plot(x_grid, y_grid, type = "l", xlab = "Test statistic", ylab = "Density", xlim = c(-3.75, 3.75))
polygon(c(x_grid[x_grid < -t_star], rev(x_grid[x_grid < -t_star])),
        c(y_grid[x_grid < -t_star], rep(0, length(y_grid[x_grid < -t_star]))), col = "skyblue")

polygon(c(x_grid[x_grid > t_star], rev(x_grid[x_grid > t_star])),
        c(y_grid[x_grid > t_star], rep(0, length(y_grid[x_grid > t_star]))), col = "skyblue")
abline(v = t_star, lty = 2)
abline(v = -t_star, lty = 2)
abline(h = 0)
```

## ANOVA


```{r}
anova(fit)
```

