---
title: "STAT 8020 R Lab 10: Multiple Linear Regression VI"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Regression with Both Quantitative and Qualitative Predictors: Salaries for Professors Data Set

The 2008-09 nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S. The data were collected as part of the on-going effort of the college's administration to monitor salary differences between male and female faculty members.

### Load and plot the data

```{r}
library(carData)
data("Salaries")
attach(Salaries) 
head(Salaries)
summary(Salaries)
boxplot(salary ~ sex, data = Salaries, las = 1)
boxplot(salary ~ discipline, data = Salaries, las = 1)
boxplot(salary ~ rank, data = Salaries, las = 1)
xtabs(~ sex + rank + discipline, data = Salaries)
plot(yrs.since.phd, salary, las = 1, pch = 16, cex = 0.75,
     xlab = "Years of service", ylab = "9-month salary")
grid()
```

### Model fitting 

```{r}
m1 <- lm(salary ~ discipline + rank + sex + yrs.since.phd, data = Salaries)
X <- model.matrix(m1)
summary(m1)

plot(m1, which = 1, cex = 0.4)

yr.range <- tapply(yrs.since.phd, list(discipline, sex, rank), range)
sex.col <- ifelse(sex == "Male", "blue", "red")
dis.col <- ifelse(discipline == "A", 16, 1)

beta0 <- m1$coefficients[1]
betaDisp <- m1$coefficients[2]
betaAssoc <- m1$coefficients[3]
betaProf <- m1$coefficients[4]
betaMale <- m1$coefficients[5]
beta1 <- m1$coefficients[6]

library(scales)
# Plot the model fits by rank
assistant <- which(rank == "AsstProf")

plot(yrs.since.phd[assistant], salary[assistant],
     pch = dis.col[assistant], cex = 0.8,
     col = alpha(sex.col[assistant], 0.5),
     yaxt = "n", xlab = "Years of service",
     main = "9-month salary", ylab = "")
axis(2, at = seq(63000, 99000, len = 6),
     labels = paste(seq(63000, 99000, len = 6)/ 1000, "k"),
     las = 1)

segments(yr.range[[1]][1], beta0 + yr.range[[1]][1] * beta1,
         yr.range[[1]][2], beta0 + yr.range[[1]][2] * beta1,
         col = "red", lwd = 1.8)
segments(yr.range[[2]][1], beta0 + betaDisp + yr.range[[2]][1] * beta1,
         yr.range[[2]][2], beta0 + betaDisp + yr.range[[2]][2] * beta1,
         col = "red", lty = 2, lwd = 1.8)
segments(yr.range[[3]][1], beta0 + betaMale + yr.range[[3]][1] * beta1,
         yr.range[[3]][2], beta0 + betaMale + yr.range[[3]][2] * beta1,
         col = "blue", lwd = 1.8)
segments(yr.range[[4]][1], beta0 + betaDisp + betaMale + yr.range[[4]][1] * beta1,
         yr.range[[4]][2], beta0 + betaDisp + betaMale + yr.range[[4]][2] * beta1,
         col = "blue", lty = 2, lwd = 1.8)

assoc <- which(rank == "AssocProf")
plot(yrs.since.phd[assoc], salary[assoc],
     pch = dis.col[assoc], cex = 0.8,
     col = alpha(sex.col[assoc], 0.5),
     yaxt = "n", xlab = "Years of service",
     main = "9-month salary", ylab = "")
axis(2, at = seq(62000, 127000, len = 6),
     labels = paste(seq(62000, 127000, len = 6)/ 1000, "k"),
     las = 1)

segments(yr.range[[5]][1], beta0 + betaAssoc + yr.range[[5]][1] * beta1,
         yr.range[[5]][2], beta0 + betaAssoc + yr.range[[5]][2] * beta1,
         col = "red", lwd = 1.8)
segments(yr.range[[6]][1], beta0 + betaDisp + betaAssoc + yr.range[[6]][1] * beta1,
         yr.range[[6]][2], beta0 + betaDisp + betaAssoc + yr.range[[6]][2] * beta1,
         col = "red", lty = 2, lwd = 1.8)
segments(yr.range[[7]][1], beta0 + betaAssoc + betaMale + yr.range[[7]][1] * beta1,
         yr.range[[7]][2], beta0 + betaAssoc + betaMale + yr.range[[7]][2] * beta1,
         col = "blue", lwd = 1.8)
segments(yr.range[[8]][1], beta0 + betaDisp + betaAssoc + betaMale + yr.range[[8]][1] * beta1,
         yr.range[[8]][2], beta0 + betaDisp + betaAssoc + betaMale + yr.range[[8]][2] * beta1,
         col = "blue", lty = 2, lwd = 1.8)

prof <- which(rank == "Prof")
plot(yrs.since.phd[prof], salary[prof],
     pch = dis.col[prof], cex = 0.8,
     col = alpha(sex.col[prof], 0.5),
     yaxt = "n", xlab = "Years of service",
     main = "9-month salary", ylab = "")
axis(2, at = seq(57000, 232000, len = 6),
     labels = paste(seq(57000, 232000, len = 6)/ 1000, "k"),
     las = 1)

segments(yr.range[[9]][1], beta0 + betaProf + yr.range[[9]][1] * beta1,
         yr.range[[9]][2], beta0 + betaProf + yr.range[[9]][2] * beta1,
         col = "red", lwd = 1.8)
segments(yr.range[[10]][1], beta0 + betaDisp + betaProf + yr.range[[10]][1] * beta1,
         yr.range[[10]][2], beta0 + betaDisp + betaProf + yr.range[[10]][2] * beta1,
         col = "red", lty = 2, lwd = 1.8)
segments(yr.range[[11]][1], beta0 + betaProf + betaMale + yr.range[[11]][1] * beta1,
         yr.range[[11]][2], beta0 + betaProf + betaMale + yr.range[[11]][2] * beta1,
         col = "blue", lwd = 1.8)
segments(yr.range[[12]][1], beta0 + betaDisp + betaProf + betaMale + yr.range[[12]][1] * beta1,
         yr.range[[12]][2], beta0 + betaDisp + betaProf + betaMale + yr.range[[12]][2] * beta1,
         col = "blue", lty = 2, lwd = 1.8)



m2 <- lm(salary ~ sex * yrs.since.phd)
summary(m2)
coeff <- m2$coefficients
plot(yrs.since.phd, salary, las = 1, pch = 16, cex = 0.5, col = alpha(sex.col, 0.5),
     xlab = "Years of service", ylab = "9-month salary")
grid()
abline(coeff[1], coeff[3], col = "red")
abline(coeff[1] + coeff[2], coeff[3] + coeff[4],
       col = "blue")
legend("toplef", legend = c("Female", "Male"),
       pch = 16, col = c("red", "blue"),
       bty = "n")

m3 <- lm(salary ~ discipline * yrs.since.phd)
summary(m3)
coeff <- m3$coefficients
plot(yrs.since.phd, salary, las = 1, pch = dis.col, cex = 0.5, col = alpha(sex.col, 0.5),
     xlab = "Years of service", ylab = "9-month salary")
grid()
abline(coeff[1], coeff[3], col = "red")
abline(coeff[1] + coeff[2], coeff[3] + coeff[4],
       col = "blue")
legend("toplef", legend = c("Female", "Male"),
       pch = 16, col = c("red", "blue"),
       bty = "n")
```

## Polynomial regression: Housing Values in Suburbs of Boston

```{r}
library(MASS)
data(Boston)

plot(Boston$lstat, Boston$medv, col = "gray", pch = 16,
     cex = 0.6, las = 1, xlab = "lower status of the population (percent)", ylab = "median value of owner-occupied homes")

m1 <- lm(medv ~ lstat, data = Boston)
abline(m1)

m2 <- lm(medv ~ lstat + I(lstat^2), data = Boston)
lines(sort(Boston$lstat), m2$fitted.values[order(Boston$lstat)], col = "red")

m3 <- lm(medv ~ lstat + I(lstat^2)+ I(lstat^3), data = Boston)
lines(sort(Boston$lstat), m3$fitted.values[order(Boston$lstat)], col = "blue")

anova(m2, m3)

m2new <- lm(medv ~ poly(lstat, 2), data = Boston)
m3new <- lm(medv ~ poly(lstat, 3), data = Boston)
summary(m3new)

anova(m2new, m3new)
```



