---
title: "STAT 8020 R Lab 5: Multiple Linear Regression I"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Species diversity on the Galapagos Islands

### First Step: Load the data

```{r}
#installinstall.packages("faraway")
library(faraway)
data(gala)

gala

#Out the data in csv
#write.csv(gala, file = "gala.csv")
```

### Plot the pairwise scatterplots

```{r}
plot(gala[, -2], cex = 0.75, col = "blue", las = 1)
```

### Correlation matrix

```{r}
cor(gala[, -2])
```

### Model 1: Fitting a simple linear regression

Here we use *Elevation* as the predictor as it has the highest correlation with *Species*

```{r}
M1 <- lm(Species ~ Elevation, data = gala)
summary(M1)
```

### Model 2: Adding *Area*

```{r}
M2 <- lm(Species ~ Elevation + Area, data = gala)
summary(M2)
```

### Model 3: Adding *Adjacent*

```{r}
M3 <- lm(Species ~ Elevation + Area + Adjacent, data = gala)
summary(M3)
```

### Full Model

```{r}
M4 <- lm(Species ~ Elevation + Area + Adjacent + Nearest + Scruz, data = gala)
summary(M4)
```

### Parameter Estimation

```{r}
X <- model.matrix(M4)
y <- gala$Species
# regression parameters
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
#beta_hat_faster <- solve(crossprod(X), crossprod(X, y))
# fitted values
y_hat <- X %*% solve(t(X) %*% X) %*% t(X) %*% y
```

### ANOVA

```{r}
anova(M4)
```

### General Linear Test

```{r}
anova(M1, M2)

par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
xg <- seq(0, 3, 0.01)
yg <- df(xg, 1, 27)
plot(xg, yg, type = "l", xaxs = "i", yaxs = "i", lwd = 1.6,
     xlab = "F test statistic", ylab = "Density")
abline(v = 0.5254, lty = 2, col = "gray")
polygon(c(xg[xg > 0.5254], rev(xg[xg > 0.5254])),
        c(yg[xg > 0.5254], rep(0, length(yg[xg > 0.5254]))),
        col = "skyblue")
```

## Simulation

### $R^2$ vs. $R^{2}_{adj}$

```{r}
set.seed(123)
N = 500
x1 <- replicate(N, rnorm(30))
x2 <- replicate(N, rnorm(30))
y1 <- apply(x1, 2, function(x) 5 + 2 * x + rnorm(30, 0, 1))
R.sq <- array(dim = c(N, 4))
for (i in 1:N){
  R.sq[i, 1] = summary(lm(y1[, i] ~ x1[, i]))$r.squared
  R.sq[i, 2] = summary(lm(y1[, i] ~ x1[, i]))$adj.r.squared
  R.sq[i, 3] = summary(lm(y1[, i] ~ x1[, i] + x2[, i]))$r.squared
  R.sq[i, 4] = summary(lm(y1[, i] ~ x1[, i] + x2[, i]))$adj.r.squared
}


par(las = 1)
plot(R.sq[, 1], R.sq[, 3], pch = 16, cex = 0.65, col = "blue",
     xlab = expression(paste("Model 1: ", R^2)),
     ylab = expression(paste("Model 2: ", R^2)))
abline(0, 1)

boxplot(R.sq[, 3] - R.sq[, 1], las = 1)
abline(h = 0, lty = 2, col = "red")

par(las = 1, mfrow = c(1, 2), mar = c(5.1, 4.6, 1.1, 1.1))
plot(R.sq[, 1], R.sq[, 3], pch = 16, cex = 0.65, col = "blue",
     xlab = expression(paste("Model 1: ", R^2)),
     ylab = expression(paste("Model 2: ", R^2)))
abline(0, 1)

plot(R.sq[, 2], R.sq[, 4], pch = 16, cex = 0.5, col = "blue",
     xlab = expression(paste("Model 1: ", R[adj]^2)),
     ylab = expression(paste("Model 2: ", R[adj]^2)))
abline(0, 1)

boxplot(R.sq[, 3] - R.sq[, 1], R.sq[, 4] - R.sq[, 2], las = 1)
abline(h = 0, lty = 2, col = "red")
axis(1, at = 1:2, labels = c("R^2", "R^2_adj"))

GLM_p <- numeric(500)

for (i in 1:500){
  reduce <- lm(y1[, i] ~ x1[, i])
  full <- lm(y1[, i] ~ x1[, i] + x2[, i])
  GLM_p[i] <- anova(reduce, full)$`Pr(>F)`[2]
}

hist(GLM_p, 30, las = 1)
```

### Multicollinearity

```{r}
library(MASS)

x <- replicate(N, mvrnorm(n = 30, c(0, 0), matrix(c(1, 0.9, 0.9, 1), 2)))
y <- array(dim = c(30, N))
for (i in 1:N){
  y[, i] = 4 + 0.8 * x[, 1, i] + 0.6 * x[, 2, i] + rnorm(30)
}
beta <- array(dim = c(3, N))
for (i in 1:N){
  beta[, i] <- lm(y[, i] ~ x[, 1, i] + x[, 2, i])$coefficients
}
```


