---
title: 'STAT 8010 R Lab 5: Binomial, Hypergeometric, and Normal Distributions '
author: "Whitney Huang"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Binomial distribution

The probability mass function of a binomial random variable with parameters $n$, the number of trials, and $p$, the success probability of every trial is
$$p_{X}(x) = \binom{n}{x}(p)^{x}(1-p)^{n-x},\, x = 0, 1, \cdots, n$$


## ESP example

To test for Extrasensory perception (ESP), we have 4 cards. They will be shuffled and one randomly selected each time,
and you are to guess which card is selected. This is repeated 10 times. Suppose you do not have
ESP. Let $R$ be the number of times you guess a card correctly. What are the distribution
and parameter(s) of $R$? What is the expected value of $R$? Furthermore, suppose that you
get certified as having ESP if you score at least an 8 on the test. What is the probability
that you get certified as having ESP?

```{r}
n = 10; p = 1 / 4
choose(10, 8)

#P(R=8)
choose(n, 8) * p^8 * (1 - p)^(n - 8) 
dbinom(8, n, p)
sum(dbinom(8:10, n, p)) # P(R=8) + P(R=9) + P(R=10)
1 - pbinom(7, n, p) # 1 - (P(R=0) + P(R=1) + ... +P(R=7))
pbinom(7, n, p, lower.tail = F) # P(R=8) + P(R=9) + P(R=10)
```

## Coke blind taste test

What is the probability that $X$ is at least 1?
```{r}
n = 4; p = .95
1 - 0.05^4
1 - dbinom(0, n, p)
```

What is the probability that $X$ is at most 3?

```{r}
sum(dbinom(0:3, n, p))
pbinom(3, n, p)
```

## General continuous random variable

```{r}
x <- seq(1, 4, 0.01)
fx <- (4 * x - x^2) / 9

plot(x, fx, type = "l", las = 1, xlim = c(0, 5))
segments(1, 0, 1, 1 / 3)
abline(h = 0)
polygon(c(x, rev(x)), c(fx, rep(0, length(fx))), col = "skyblue")
abline(v = 2.25, lwd = 2, col = "red")
axis(3, at = 2.25, labels = "Mean", line = 0, col = "red",
     lwd = 2)
axis(1, at = 2.25, line = -1, col = "red")
abline(v = 1.75, lty = 2); abline(v = 2.75, lty = 2)
```

## Normal distribution

1. $\Phi(0)=.50\Rightarrow$} Mean and Median ($50_{th}$ percentile) for standard normal are both $0$

2. $\Phi(-z)=1-\Phi(z)$}

3. $\mathbb{P}(Z>z)=1-\Phi(z)=\Phi(-z)$}

```{r}
pnorm(0)

pnorm(-1)

1 - pnorm(1)

pnorm(1, lower.tail = F)

pnorm(-1.75)
pnorm(2) - pnorm(-2)
```


