---
title: 'STAT 8010 R Lab 1: Read data into R'
author: "Whitney Huang"
date: "8/20/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab Objective

* To gain experience with R, a programming language and free software environment for statistical computing and graphics.
* To read data into R.

### Setup

* You should have R installed, if not, open a web browser and go to (http://cran.r-project.org) and download and install R. It also helpful to install RStudo (http://rstudio.com).

* Create a folder for this R lab. Download the Maximum Heart Rate dataset at (http://whitneyhuang83.github.io/maxHeartRate.csv) and save it in the folder you just created.

### Load a max heart rate dataset

There are several ways to load a dataset into R:

+ Importing Data over the Internet

```{r}
dat <-read.csv('http://whitneyhuang83.github.io/STAT8010/Data/maxHeartRate.csv', header = T) 
```

Let's take a look at the data

```{r}
dat
```

+ Read the dataset from you computer

```{r, message=FALSE}
dat <- read.csv('maxHeartRate.csv', header = T) 
```

+ If the data is not too big, you can type the data into R

```{r}
age <- c(18, 23, 25, 35, 65, 54, 34, 56, 72, 19, 23, 42, 18, 39, 37)
maxHeartRate <- c(202, 186, 187, 180, 156, 169, 174, 172, 153,
                  199, 193, 174, 198, 183, 178)
dat <- data.frame(cbind(age, maxHeartRate))
```

### Load a Facebook dataset

```{r}
FBdata <- read.csv('/Users/wkhuang/Desktop/Desktop - mass-mini19-huang/Teaching/R/FacebookFriends.csv',
                   header = T) 
str(FBdata)
```


### Loading a built-in R data

```{r}
data("mtcars")
head(mtcars, 6)
?mtcars
```




