---
title: "STAT 8010 R Lab 17: Simple Linear Regression III"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Understanding Sampling Distributions and Confident Intervals via simulation

Simulate the "data" $\{x_{i}, y_{i}\}_{i=1}^{n}$ where $y_{i} = \beta_{0}+\beta_{1}x_{i} + \varepsilon_{i}$, $\varepsilon \sim \mathrm{N}(0, \sigma^2)$. Repeat this process $N$ times.

### Generate data in R

```{r}
set.seed(12)
n = 30; beta0 = 3; beta1 = 1.5; N = 100; sigma2 = 1 
x <- 3 + 5 * runif(n)
set.seed(123)
y <- replicate(N, beta0 + beta1 * x + rnorm(n, mean = 0, sd = sqrt(sigma2)))
dim(y)
```

### Plot the first few simulated datasets

```{r}
for (i in 1:5){
  plot(x, y[, i], pch = 16, las = 1, ylab = "y")
  abline(3, 1.5, col = "blue", lwd = 1.5)
}
```

### Estimate the $\beta_{0}$, $\beta_{1}$, and $\sigma^2$ for each simulated dataset 

```{r}
beta0_hat <- beta1_hat <- sigma2_hat <- se_beta1 <- numeric(N)
for (i in 1:100){
  fit <- lm(lm(y[, i] ~ x))
  beta0_hat[i] <- summary(fit)[["coefficients"]][, 1][1]
  beta1_hat[i] <- summary(fit)[["coefficients"]][, 1][2]
  se_beta1[i] <- summary(fit)[["coefficients"]][, 2][2]
  sigma2_hat[i] <- summary(fit)[["sigma"]]^2
}
```

### Assess the estimation perfromance 

```{r}
boxplot(beta0_hat, las = 1, main = expression(hat(beta[0])))
abline(h = beta0, col = "blue", lwd = 1.5)

boxplot(beta1_hat, las = 1, main = expression(hat(beta[1])))
abline(h = beta1, col = "blue", lwd = 1.5)

boxplot(sigma2_hat, las = 1, main = expression(paste("Boxplot of ", hat(sigma)^2)))
abline(h = sigma2, col = "blue", lwd = 1.5)
```

### Sampling distribution

```{r}
hist(beta1_hat, 16, col = "lightblue", border = "gray",
     main = expression(paste("Histogram of ", hat(beta)[1])),
     xlab = expression(hat(beta)[1]))
abline(v = beta1, col = "blue", lwd = 1.5)
mtext(expression(beta[1]), 1, at = beta1, col = "blue")
```

### CI's for all the simulated datasets

```{r}
t <- qt(1 - 0.05 / 2, n - 2)
LL <- beta1_hat - t * se_beta1
UL <- beta1_hat + t * se_beta1
miss <- which((LL - beta1) * (UL - beta1) > 0)


par(las = 1)
plot(1:100, rep(beta1, N), type = "l", bty = "n", xaxt = "n", xlab = "",
     lwd = 1.5, ylab = expression(hat(beta)[1]))
for (i in 1:100){
  segments(i, LL[i], i, UL[i], col = "blue")
}

for (i in miss){
  segments(i, LL[i], i, UL[i], col = "red")
}
```


## ANOVA

### First Step: Load the data

```{r}
dat <- read.csv('http://whitneyhuang83.github.io/STAT8010/Data/maxHeartRate.csv', header = T) 
head(dat)
attach(dat)
```

### Fitting a simple linear regression

```{r}
fit <- lm(MaxHeartRate ~ Age)
summary(fit)
R.sq <- summary(fit)[["r.squared"]]
r <- cor(dat$Age, dat$MaxHeartRate)
r^2; R.sq
```

### ANOVA

```{r}
anova(fit)
```

