---
title: "STAT 8020 R Lab 13: Inference for Proportions"
author: "Whitney"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cancer treatment survival rate

Researchers in the development of new treatments for cancer patients often
evaluate the effectiveness of new therapies by reporting the *proportion* of patients who survive for a specified period of time after completion of the treatment. A new genetic treatment of 870 patients with a particular type of cancer resulted in 330 patients surviving at least 5 years after treatment. *Estimate* the proportion of all patients with the specified type of cancer who would survive at least 5 years after being administered this treatment.

```{r, message=FALSE}
n = 870; x = 330 
# point estimate
phat <- x / n
phat
# 95% CI for p
library(fastR)
wald.ci(x, n, conf.level = 0.95)
# 99% CI for p
wald.ci(x, n, conf.level = 0.9)
```

## Bird flu example

Among 900 randomly selected registered
voters nationwide, 63% of them are somewhat or
very concerned about the spread of bird flu in
the United States. 

* What is the point estimate for $p$, the proportion of U.S. voters who are concerned about the spread of bird flu?

* Construct a 95\% CI for $p$


```{r}
n = 900; x = 900 * .63 
# point estimate
phat <- x / n
phat
# 95% CI for p
wald.ci(x, n, conf.level = 0.95)
```

## Sample size calculation example

A researcher wants to estimate the proportion of voters who will vote for candidate A. She wants to estimate to within 0.05 with 90% confidence.

```{r}
# True proportion is .9
p = 0.9; alpha = 1 - 0.9
n = p * (1 - p) * (qnorm(1 - alpha / 2) / 0.05)^2
n
# True proportion is .6
p = 0.6
n = p * (1 - p) * (qnorm(1 - alpha / 2) / 0.05)^2
n
# True proportion is .5
p = 0.5
n = p * (1 - p) * (qnorm(1 - alpha / 2) / 0.05)^2
n
# Just for fun
p <- seq(0.01, 0.99, 0.01)
n = p * (1 - p) * (qnorm(1 - alpha / 2) / 0.05)^2

plot(p, n, type = "l",
     xlab = "True proportion",
     ylab = "Simple size",
     las = 1)
abline(v = 0.5, col = "gray")
abline(h = max(n), col = "gray")
grid()
```

## Hypothesis testing: Bird flu example

Among 900 randomly selected registered
voters nationwide, 63% of them are somewhat or
very concerned about the spread of bird flu in the United States. Conduct a hypothesis test at .01 level to assess the research hypothesis: $p>.6$.


```{r}
n = 900; x = .63 * 900
p_null = .6; alpha = 0.01 

prop.test(x, n, p = p_null,
          conf.level = 1 - alpha,
          alternative = "greater",
          correct = FALSE)
# With Yates' continuity correction
prop.test(x, n, p = p_null,
          conf.level = 1 - alpha,
          alternative = "greater",
          correct = TRUE)
```

## Proportion of CU vegetarian

```{r}
n = 25; x = 0
wald.ci(x, n)
## Wilson score CI
library(PropCIs)
scoreci(x, n, conf.level = .95)
## Check 
prop.test(x, n, conf.level = 0.95,
          correct = F)
```

Here the rule of three provides a quick and reasonable approximation

## Inference for $p_{1}-p_{2}$

A Simple Random Simple of 100 CU graduate students is taken and it is found that 79 strongly
agree that they would recommend their current graduate program. A Simple Random Simple of 85
USC graduate students is taken and it is found that 52 strongly agree that they
would recommend their current graduate program. At 5 % level, can we conclude that the proportion of strongly agree is higher at CU?


```{r}
x <- c(79, 52); n <- c(100, 85)
prop.test(x, n, alternative = "greater", correct = F)
# With Yates' continuity correction
prop.test(x, n, alternative = "greater", correct = F)
```



