---
title: 'STAT 8010 R Lab 4: Data Summary/Visualization III'
author: "Whitney Huang"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Interquartile range (IQR)

```{r}
data1 <- c(13, 18, 13, 14, 13, 16, 14, 21, 13)
IQR(data1, type = 1)
data2 <- c(13, 18, 13, 14, 13, 16, 14, 210, 13)
IQR(data2, type = 1)
```

### Percentiles

```{r}
#Q1
quantile(data1, 0.25, type = 1)
#Q2 aka median
quantile(data1, 0.5, type = 1)
#Q3
quantile(data1, 0.75, type = 1)


data3 <- c(13, 18, 13, 14, 13, 16, 14, 21, 13, 9,
           27, 18, 25, 20, 6)
quantile(data3, c(0.35, 0.65), type = 1)
```

### Boxplot

```{r}
boxplot(data1)
boxplot(data1, horizontal = T)
```

### Load the ORD flight dataset

```{r}
url <- "https://whitneyhuang83.github.io/STAT8010/Data/flights.csv"
ORD <- read.csv(url, header = TRUE)
```

### Let's take a look at the data

```{r}
dim(ORD)
n <- dim(ORD)[1]
head(ORD)
```

### 2 way Frequency Table

```{r}
tab3 <- table(ORD[, c("carrier", "origin")])
tab3

tab4 <- table(ORD[, c("carrier", "origin")])/n
tab4
```

### Stacked/dodged bar chart

```{r}
## Stacked bar chart
barplot(tab3, xlab = "Origin", col = c("darkblue","red"), args.legend = list(x = "topleft"), las = 1)
legend("topleft", legend = c("UA", "AA"),
       pch = 15, col = c("red", "blue"), bty = "n", cex = 1.25, title = "Carrier")

## Dodged bar chart
barplot(tab3, xlab = "Origin", col = c("darkblue","red"), args.legend = list(x = "topleft"), las = 1, beside = T)
legend("topleft", legend = c("UA", "AA"),
       pch = 15, col = c("red", "blue"), bty = "n", cex = 1.25, title = "Carrier")
```

### Qualitative vs Quantitative: Side by Side Boxplots

```{r}
attach(ORD)
library(tidyverse)
boxplot(arr_delay ~ carrier, filter(ORD, arr_delay > 10), boxwex = 0.35,
        col = c("blue", "red"),
        staplewex = 0.35, outwex = 0.35,
        cex.axis = 1.5, las = 1, log = "y",
        outcol = c("blue", "red"),
        outcex = 0.35, main = "Arrival Delay vs. Carrier")
abline(h = 11, lty = 2, col = "gray")
```

### Quantitative vs Quantitative: Scatter Plot

```{r}
url <- "https://whitneyhuang83.github.io/STAT8010/Data/maxHeartRate.csv"
dat <- read.csv(url, header = TRUE)

par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(dat$Age, dat$MaxHeartRate, pch = 16, xlab = "Age", ylab = "Max heart rate (bpm)")
grid()
```


### Visualizing Time Series Data: Mauna Loa Atmospheric CO2 Concentration

Atmospheric concentrations of $\text{CO}_{2}$ are expressed in parts per million (ppm) and reported in the preliminary 1997 SIO manometric mole fraction scale.


```{r}
data("co2")
par(las = 1)
ts.plot(co2, ylab = expression(paste(CO[2], " (ppm)")))

time <- seq(1959, 1998, len = 468)
plot(time[1:36], co2[1:36], type = "l", xlab = "Year",
     ylab = expression(paste(CO[2], " (ppm)")),
     las = 1)

plot(time[445:468], co2[445:468], type = "l", xlab = "Year",
     ylab = expression(paste(CO[2], " (ppm)")),
     las = 1)
```

### Visualizing  Cross-Sectional Data

```{r}
library(maps)
library(ggmap)
data("USArrests")
USArrests$region <- tolower(row.names(USArrests))

statesMap <- map_data("state")
str(statesMap)

murderMap <- merge(statesMap, USArrests, by = "region")
str(murderMap)

ggplot(murderMap, aes(x = long, y = lat, group = group, fill = Murder)) + geom_polygon(color = "black") + scale_fill_gradient(low = "lightblue", high = "red", guide = "legend")
```

### Visualizing Spatio-Temporal Data: ERA-Interim

The ERA-Interim is a global atmospheric reanalysis dataset. Reanalysis is an approach to
produce spatially and temporally gridded datasets via data assimilation for climate monitoring and analysis.

```{r}
load("ERA_tmx_2010_JanFeb.RData")
library(fields)
par(mar = c(4.6, 4.1, 2.1, 0))
for (i in seq(1:5)){
        image.plot(lon, lat, tmx_dat[,, i], las = 1, main = format(day[i], "%m/%d/%Y"))
        map("state", xlim = range(lon), ylim = range(lat),
            add = T)
        map("world", xlim = range(lon), ylim = range(lat),
            add = T)      
}
```


```{r, eval=FALSE}
library(animation)
saveLatex({
  for (i in 1:58){
          image.plot(lon, lat, tmx_dat[,, i], las = 1, main = format(day[i], "%m/%d/%Y"),
                     zlim = range(tmx_dat))
          map("state", xlim = range(lon), ylim = range(lat), add = T)
          map("world", xlim = range(lon), ylim = range(lat), add = T)
  }
}, img.name = "ERA_Tmax", ani.opts = "controls,width=0.975\\textwidth",
latex.filename = ifelse(interactive(), "ERA_TMX_JanFeb.tex", ""),
interval = 0.5, nmax = 58, ani.dev = "pdf", ani.type = "pdf", ani.width = 8,
ani.height = 6, documentclass = paste("\\documentclass{article}",
                                     "\\usepackage[papersize={8in,6in},margin=0.1in]{geometry}",
                                     sep = "\n"))
```









