---
title: "STAT 8010 R Lab 9: Hypothesis Testing"
author: "Whitney Huang"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Cereal Weight Example

New Age Granola Inc claims that average weight of its
cereal boxes is 16 oz. The Genuine Grocery Corporation
will send back a shipment if the average weight is any
less.
Suppose Genuine Grocery Corporation takes a random
sample of 49 boxes, weight each one, and compute
the sample mean $\bar{X}=15.9$ oz and sample standard
deviation $s = 0.35$ oz.
Perform a hypothesis test at 0.05 significant level to determine
if they would reject $H_{0}$, and therefore, this shipment

1. $H_{0}: \mu = 16$ vs. $H_{a}: \mu <16$

2. Test Statistic: $t_{obs} = \frac{15.9-16}{0.35/\sqrt{49}} = -2$

3. Rejection Region Method: $-t_{0.05, 48} = -1.68 \Rightarrow$ Rejection Region is $(-\infty, -1.68]$. Since $t_{obs}$ is in rejection region, we reject $H_{0}$  

4. P-Value Method: $\mathbb{P}(t^{*} \le -2) = 0.0256 < \alpha = 0.05 \Rightarrow$ reject $H_{0}$ 

5. Draw a Conclusion: We do have enough statistical evidence to conclude that the average weight is less than 16 oz at 0.05 significant level 

```{r}
# P-value
pt(-2, 48)

curve(dt(x, 48), from = -3.5, to = 3.5, n = 1001, lwd = 1.5, col = "blue",
      xlab = "Test statistic",
      ylab = "Density", las = 1)
abline(v = -2)
x_grid <- seq(-3.5, 3.5, 0.01)
y_grid <- dt(x_grid, 48)
polygon(c(x_grid[x_grid < -2], rev(x_grid[x_grid < -2])),
        c(y_grid[x_grid < -2], rep(0, length(y_grid[x_grid < -2]))), col = "pink")
abline(h = 0, lty = 3, col = "gray")
axis(1, at = -2, labels = expression(t["obs"]),
     line = -0.75)
abline(v = -qt(0.95, 48), lty = 2, col = "red")
arrows(-1.68, 0.25, x1 = -3.5, length = 0.1,
       lty = 2, col = "red")
text(-3, 0.3, "Rejection Region", col = "red")
```

## Blood Test Example

A series of blood tests were run on a particular patient over five
days. It is of interest to determine if the mean blood protein for
this patient differs from 7.25, the value for healthy adults.
Suppose the sample mean ($n=20$) is 7.35 and sample standard
deviation is 0.5. Perform a hypothesis test using significance
level of 0.05


1. $H_{0}: \mu = 7.25$ vs. $H_{a}: \mu \neq 7.25$

2. $t_{obs} = \frac{7.35-7.25}{0.5/\sqrt{20}} = 0.8944$

3. P-value: $2 \times \mathbb{P}(t^{*} \ge 0.8944) = 0.3823 > 0.05$

4. We do not have enough statistical evidence to conclude that the mean blood protein is different from 7.25 at 5% significant level

```{r}
curve(dt(x, 19), from = -3.75, to = 3.75, n = 1001, lwd = 1.5, col = "blue",
      xlab = "Test statistic",
      ylab = "Density", las = 1)
x_grid <- seq(-3.75, 3.75, 0.01)
y_grid <- dt(x_grid, 19)
t_star = 0.8944
polygon(c(x_grid[x_grid < -t_star], rev(x_grid[x_grid < -t_star])),
        c(y_grid[x_grid < -t_star], rep(0, length(y_grid[x_grid < -t_star]))), col = "pink")
polygon(c(x_grid[x_grid > t_star], rev(x_grid[x_grid > t_star])),
        c(y_grid[x_grid > t_star], rep(0, length(y_grid[x_grid > t_star]))), col = "pink")
abline(h = 0, lty = 3, col = "gray")
axis(1, at = 0.8944, labels = expression(t["obs"]), line = -0.75)
axis(1, at = -0.8944, labels = expression(-t["obs"]), line = -0.75)
abline(v = -qt(0.975, 19), lty = 2, col = "red")
abline(v = qt(0.975, 19), lty = 2, col = "red")
abline(v = 0.8944)
arrows(-2.09, 0.25, x1 = -3.75, length = 0.1,
       lty = 2, col = "red")
arrows(2.09, 0.25, x1 = 3.75, length = 0.1,
       lty = 2, col = "red")
text(-3, 0.3, "Rejection Region", col = "red")
text(3, 0.3, "Rejection Region", col = "red")
```

## Type I and Type II Errors

```{r}
library(scales)
curve(dnorm(x), from = -3.75, to = 6.75, n = 1001, lwd = 1.5, col = "gray",
      xlab = "",
      ylab = "", las = 1,
      xaxt = "n", yaxt = "n")
abline(v = 1.96, lty = 2, col = "blue")
x_grid <- seq(-3.75, 6.75, 0.01)
y_grid <- dnorm(x_grid)
polygon(c(x_grid[x_grid > 1.96], rev(x_grid[x_grid > 1.96])),
        c(y_grid[x_grid > 1.96], rep(0, length(y_grid[x_grid > 1.96]))), col = alpha("pink", 0.2))
curve(dnorm(x, mean = 3), from = -3.75, to = 6.75, n = 1001, lwd = 1.5, add = T,
      xlab = "",
      ylab = "", las = 1)
x_grid <- seq(-3.75, 6.75, 0.01)
y_grid <- dnorm(x_grid, mean = 3)
polygon(c(x_grid[x_grid < 1.96], rev(x_grid[x_grid < 1.96])),
        c(y_grid[x_grid < 1.96], rep(0, length(y_grid[x_grid < 1.96]))), col = alpha("lightblue", 0.2))
abline(v = 0, lty = 2, col = "gray")
abline(v = 3, lty = 2)
axis(1, at = 0, labels = expression(mu[0]))
axis(1, at = 3, labels = expression(mu[a]))
text(2.2, 0.02, expression(alpha), col = "red")
text(1.6, 0.02, expression(beta), col = "blue")
```


