---
title: 'STAT 8010 R Lab 2: Data Summary/Visualization I'
author: "Whitney Huang"
date: "8/20/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load the dataset

There are several ways to load a dataset into R:

+ Importing data over the Internet

```{r}
sport <- read.table("https://whitneyhuang83.github.io/STAT8010/Data/sport.txt", header = TRUE)
```

Let's take a look at the data

```{r}
#sport
head(sport) # print the first 6 observations
```

+ Read the dataset from you computer

```{r}
# Set working directory
setwd("/Users/wkhuang/Desktop/Desktop - mass-mini19-huang/Teaching/R/20Fall")
# This is the path  of the folder (in your computer). 
getwd()
dir()
sport1 <- read.table("sport.txt", header = TRUE)
```

### Frequency Table

```{r}
tab1 <- table(sport)
tab1 # print the table
# Relative frequency 
n <- dim(sport)[1] # sample size
tab2 <- table(sport)/n
tab2
```

### Bar Chart

```{r}
# Bart chart for the frequency
par(las = 2, mar = c(7.1, 4.1, 1.1, 1.1))
barplot(tab1, ylab = "Frequency")
# Bart chart for the relative frequency
par(las = 2, mar = c(7.1, 4.1, 1.1, 1.1))
barplot(tab2, ylab = "Relative Frequency")
# Pareto chart
par(las = 2, mar = c(7.1, 4.1, 1.1, 1.1))
barplot(sort(tab1, decreasing = T), ylab = "Frequency")
```

### Pie Chart

```{r}
par(mar = c(1.1, 3.1, 1.1, 3.1))
pie(tab1, col = rainbow(8))
# rotate the pie
par(mar = c(1.1, 3.1, 1.1, 3.1))
pie(table(sport), col = rainbow(8), init.angle = 135)
```

### Load the ORD flight dataset

```{r}
url <- "https://whitneyhuang83.github.io/STAT8010/Data/flights.csv"
ORD <- read.csv(url, header = TRUE)
```

### Let's take a look at the data

```{r}
dim(ORD)
n <- dim(ORD)[1]
head(ORD)
```

### 2 way Frequency Table

```{r}
tab3 <- table(ORD[, c("carrier", "origin")])
tab3

tab4 <- table(ORD[, c("carrier", "origin")])/n
tab4
```

### Stacked/dodged bar chart

```{r}
## Stacked bar chart
barplot(tab3, xlab = "Origin", col = c("darkblue","red"), args.legend = list(x = "topleft"), las = 1)
legend("topleft", legend = c("UA", "AA"),
       pch = 15, col = c("red", "blue"), bty = "n", cex = 1.25, title = "Carrier")

## Dodged bar chart
barplot(tab3, xlab = "Origin", col = c("darkblue","red"), args.legend = list(x = "topleft"), las = 1, beside = T)
legend("topleft", legend = c("UA", "AA"),
       pch = 15, col = c("red", "blue"), bty = "n", cex = 1.25, title = "Carrier")
```

#### Violent Crime Rates by US State

This data set contains statistics, in arrests per 100,000 residents for assault, murder, and rape in each of the 50 US states in 1973. Also given is the percent of the population living in urban areas.


```{r}
data(USArrests) # this is a bulit-in data in R
dim(USArrests)
head(USArrests)
```

### Stem-and-Leaf Plot

```{r}
stem(USArrests$Murder)
stem(USArrests$Murder, scale = 2)
```

### Histogram

```{r}
par(las = 1)
hist(USArrests$Murder, main = "Histogram of US Murder Rate in 1973",
     col = "lightblue", border = "gray", xlab = "Murder Rate (per 100,000)")
# Let's change the bin size
par(las = 1)
hist(USArrests$Murder, nclass = 15,
     main = "Histogram of US Murder Rate in 1973", col = "lightblue",
     border = "gray", xlab = "Murder Rate (per 100,000)")
# Let's change the bin size again 
par(las = 1)
hist(USArrests$Murder, nclass = 2,
     main = "Histogram of US Murder Rate in 1973", col = "lightblue",
     border = "gray", xlab = "Murder Rate (per 100,000)")
```


### Boxplot

```{r}
# Horizontal boxplot
par(las = 1)
boxplot(USArrests$Murder, main = "Murder Rate (per 100,000)", horizontal = T)
# Vertical boxplot
par(las = 1)
boxplot(USArrests$Murder, main = "Murder Rate (per 100,000)")
```


### Qualitative vs Quantitative: Side by Side Boxplots

```{r, eval=FALSE}
attach(ORD)
library(tidyverse)
boxplot(arr_delay ~ carrier, filter(ORD, arr_delay > 10), boxwex = 0.35,
        col = c("blue", "red"),
        staplewex = 0.35, outwex = 0.35,
        cex.axis = 1.5, las = 1, log = "y",
        outcol = c("blue", "red"),
        outcex = 0.35, main = "Arrival Delay vs. Carrier")
abline(h = 11, lty = 2, col = "gray")
```

### Quantitative vs Quantitative: Scatter Plot

```{r}
url <- "https://whitneyhuang83.github.io/STAT8010/Data/maxHeartRate.csv"
dat <- read.csv(url, header = TRUE)

par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(dat$Age, dat$MaxHeartRate, pch = 16, xlab = "Age", ylab = "Max heart rate (bpm)")
grid()
```












