---
title: 'STAT 8010 R Lab 6: Normal Distributions '
author: "Whitney Huang"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 8.5
    fig_height: 6.5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Normal distribution

### Normal density curve

```{r}
xg <- seq(-3.5, 3.5, 0.01)
yg <- dnorm(xg)

par(las = 1)
plot(xg, yg, type = "l", xlab = "",
     ylab = "Probability Density",
     col = "blue", xaxt = "n",
     yaxt = "n", yaxs = "i",
     ylim = c(0, 0.45))
abline(v = 0, lty = 2)
axis(1, at = 0, label = expression(mu),
     tick = F, line = -1)
axis(1, at = 0, label = "median", line = 0,
     tick = F)
axis(1, at = 0, label = "mode", line = 1,
     tick = F)
axis(1, at = -3.5, labels = expression(paste("curve extends to ", -infinity)), cex.axis = 0.7)
axis(1, at = 3.5, labels = expression(paste("curve extends to ", infinity)), cex.axis = 0.7)
segments(0, dnorm(1), -1, col = "gray")
segments(0, dnorm(1), 1, col = "gray")
segments(-1, dnorm(1), -1, 0, lty = 2,
         col = "gray")
segments(1, dnorm(1), 1, 0, lty = 2,
         col = "gray")
axis(1, at = -1, label = expression(mu - sigma), tick = F, cex.axis = 0.8, line = -1)
axis(1, at = 1, label = expression(mu + sigma), tick = F, cex.axis = 0.8, line = -1)
```

### Standard Normal: $\mathrm{Z} ~ \mathrm{N}(0, 1)$

We use $\Phi(\cdot)$ to denote the cdf of the standard normal distribution

1. $\Phi(0)=.50\Rightarrow$} *Mean* and *Median* ($50_{th}$ percentile) for standard normal are both $0$

2. $\Phi(-z)=1-\Phi(z)$}

3. $\mathbb{P}(Z>z)=1-\Phi(z)=\Phi(-z)$}

```{r}
pnorm(0)
pnorm(-1)
1 - pnorm(1)
pnorm(1, lower.tail = F)

pnorm(-1.75)
pnorm(2) - pnorm(-2)
pnorm(0.5)
```

### STAT 8020 exam score example

```{r}
mu = 78; sigma2 = 36
1 - pnorm(84, mu, sqrt(sigma2))

(pnorm(84, mu, sqrt(sigma2)) - pnorm(75, mu, sqrt(sigma2))) / (1 - pnorm(75, mu, sqrt(sigma2)))
```

### Standard normal percentiles

```{r}
qnorm(c(.1, .55, .9))
```

## General normal percentiles

```{r}
qnorm(.8, 20, 7)
20 + 7 * qnorm(.8)
```

## Normal approximation of Binomial Distribution

```{r}
n = 400; p = 0.93
sum(dbinom(370:373, n, p))
pbinom(373, n, p) - pbinom(369, n, p)

mu = n * p; sigma = sqrt(n * p * (1 - p)) 
pnorm(373.5, mu, sigma) - pnorm(369.5, mu, sigma)

plot(0:400, dbinom(0:400, n, p), type = "h", las = 1,
     xlim = c(344, 400), xlab = "x", ylab = "", col = "blue")
xg <- seq(344, 400, 0.1); yg <- dnorm(xg, mu, sigma)
lines(xg, yg, col = "blue", lty = 2)
abline(v = c(369.5, 373.5), lty = 2, col = "red")
lines(370:373, dbinom(370:373, n, p), type = "h", col = "red")
```

