---
title: 'DSA 8070 R Lab 1'
author: "your name"
date: 'August 22, 2021'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```


## Motor Trend Car Road Tests


The data was extracted from the 1974 Motor *Trend US* magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles.
You can type `r colorize("? mtcars", "blue")` to learn more about this data set. 

### Load the dataset

**Code:**

```{r}
data(mtcars)
```

### Let's take a look at the data

**Code:**

```{r}
head(mtcars)
```

In this lab, we will only focus on the following variables:

1. mpg:	Miles/gallon
2. disp: Displacement (cu.in.)
3. hp:	Gross horsepower
4. drat:	Rear axle ratio
5. wt:	Weight (1000 lbs)

**Code:**

```{r}
vars <- which(names(mtcars) %in% c("mpg", "disp", "hp", "drat", "wt"))
```

### Plot the scatterplot matrix

**Code:**

```{r}
par(las = 1)
pairs(mtcars[, vars], cex = 0.8, col = "purple")
```

### Plot the scatterplot matrix using *scatterplotMatrix* function from the *car* library 

**Code:**

```{r}

```

### Plot the scatterplot matrix using *ggpairs* function from the *GGally* library

**Code:**

```{r}

```

## Multivariate Normal Distributions

In this question, we are going to simulate data from bi-variate normal distribution with different correlation coefficients:

### $\rho=0$, that is, these two variables are indepdendent to each other

**Code:**

```{r, message=FALSE}
library(MASS)
rho <- 0
sigma <- 1
x1 <- mvrnorm(n = 1000, mu = c(0, 0),
              Sigma = matrix(c(sigma, sigma * rho, sigma * rho, sigma), 2))
library(GGally)
ggpairs(data.frame(x1))
```

### $\rho=0.4$

**Code:**

```{r}
rho <- 0.4

```

### $\rho=0.9$

**Code:**

```{r}
rho <- 0.9

```





